---
title: "Introduction to RxODE"
author: "Matthew L. Fidler, Melissa Hallow, and Wenping Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_width: 850px
fig_height: 250px
vignette: >
  %\VignetteIndexEntry{Introduction to RxODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
```

# Introduction
`RxODE` is an R package that facilitates simulation with ODE models in
R. It is designed with pharmacometrics models in mind, but can be
applied more generally to any ODE model.

# Description of RxODE illustrated through an example

The model equations can be specified through a text string, a model
file or an R expression. Both differential and algebraic equations are
permitted. Differential equations are specified by `d/dt(var_name) = `. Each
equation can be separated by a semicolon.

To load `RxODE` package and compile the model: 

```{r}
library(RxODE)
mod1 <-RxODE({
    C2 = centr/V2;
    C3 = peri/V3;
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
});
```

## Specify ODE parameters and initial conditions

Model parameters can be defined as named vectors. Names of parameters in
the vector must be a superset of parameters in the ODE model, and the
order of parameters within the vector is not important. 
```{r}
theta <- 
   c(KA=2.94E-01, CL=1.86E+01, V2=4.02E+01, # central 
     Q=1.05E+01,  V3=2.97E+02,              # peripheral
     Kin=1, Kout=1, EC50=200)               # effects
```

Initial conditions (ICs) are defined through a vector as well.  If the
elements are not specified, the initial condition for the compartment
is assumed to be zero.


```{r}
inits <- c(eff=1);
```

## Specify Dosing and sampling in RxODE

`RxODE` provides a simple and very flexible way to specify dosing and
sampling through functions that generate an event table. First, an
empty event table is generated through the "eventTable()" function:

```{r}
ev <- eventTable(amount.units='mg', time.units='hours')
```

Next, use the `add.dosing()` and `add.sampling()` functions of the
`EventTable` object to specify the dosing (amounts, frequency and/or
times, etc.) and observation times at which to sample the state of the
system.  These functions can be called multiple times to specify more
complex dosing or sampling regiments.  Here, these functions are used
to specify 10mg BID dosing for 5 days, followed by 20mg QD dosing for
5 days:

```{r}
ev$add.dosing(dose=10000, nbr.doses=10, dosing.interval=12)
ev$add.dosing(dose=20000, nbr.doses=5, start.time=120, dosing.interval=24)
ev$add.sampling(0:240)
```

If you wish you can also do this with the `mattigr` pipe operator `%>%`

```{r}
ev <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=10, dosing.interval=12) %>%
    add.dosing(dose=20000, nbr.doses=5, start.time=120,dosing.interval=24) %>%
    add.sampling(0:240);
```

The functions `get.dosing()` and `get.sampling()` can be used to
retrieve information from the event table.

```{r}
knitr::kable(head(ev$get.dosing()))
```

```{r}
knitr::kable(head(ev$get.sampling()))
```

# Solving ODEs

The ODE can now be solved by calling the model object's `run` or `solve`
function. Simulation results for all variables in the model are stored
in the output matrix x. 

```{r}
x <- mod1$solve(theta, ev, inits)
knitr::kable(head(x))
```

This returns a matrix.  You can see the compartment values in the plot
below:

```{r}
par(mfrow=c(1,2))
matplot(x[,"C2"], type="l", ylab="Central Concentration")
matplot(x[,"eff"], type="l", ylab = "Effect")
```

You can also solve this and create a RxODE data frame:

```{r, results="asis"}
x <- solve(mod1,theta, ev, inits);
rxHtml(x)
```
