---
title: "RxODE Benchmarking"
author: "Matthew L. Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RxODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Setup the RxODE models and event table for comparison

```{r}
library(RxODE)

et <- eventTable() %>% add.dosing(dose=3, nbr.doses=6, dosing.interval=8) %>%
    add.sampling(seq(0, 48, length.out=200))

ode.1c <- RxODE({
    C2 = center/V;
    d/dt(center) = - CL*C2
})

solve.1c <- RxODE({
    C2 = linCmt(CL, V);
})

ode.1c.ka <- RxODE({
    C2 = center/V;
    d/dt(depot) = -KA * depot
    d/dt(center) = KA * depot - CL*C2
})


solve.1c.ka <- RxODE({
    C2 = linCmt(V, CL, KA);
})

ode.2c <- RxODE({
    C2 = centr/V;
    C3 = peri/V2;
    d/dt(centr) = - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  = Q*C2 - Q*C3;
})

solve.2c <- RxODE({
    C2=linCmt(V, CL, V2, Q);
})

ode.2c.ka <- RxODE({
   C2 = centr/V;
   C3 = peri/V2;
   d/dt(depot) =-KA*depot;
   d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
   d/dt(peri)  =                    Q*C2 - Q*C3;
})

solve.2c.ka <- RxODE({
    C2=linCmt(V, CL, V2, Q, KA);
})


ode.3c <- RxODE({
     C2 = centr/V;
     C3 = peri/V2;
     C4 = peri2 / V3
     d/dt(centr) = - CL*C2 - Q*C2 + Q*C3  - Q2*C2 + Q2*C4;
     d/dt(peri)  = Q*C2 - Q*C3;
     d / dt(peri2) = Q2 * C2 - Q2 * C4
 })
solve.3c <- RxODE({
     ## double solvedC(double t, int parameterization, int cmt, unsigned int col, double p1, double p2, double p3, double p4, double p5, double p6, double p7, double p8);
     C2=linCmt(V, CL, V2, Q, Q2, V3);
 })
 
 
 ode.3c.ka <- RxODE({
     C2 = centr/V;
     C3 = peri/V2;
     C4 = peri2 / V3
	 d/dt(depot) =-KA*depot;
     d/dt(centr) = KA*depot- CL*C2 - Q*C2 + Q*C3  - Q2*C2 + Q2*C4;
     d/dt(peri)  = Q*C2 - Q*C3;
     d / dt(peri2) = Q2 * C2 - Q2 * C4
 })
 
solve.3c.ka <- RxODE({
     ## double solvedC(double t, int parameterization, int cmt, unsigned int col, double p1, double p2, double p3, double p4, double p5, double p6, double p7, double p8);
     C2=linCmt(V, CL, V2, Q, Q2, V3, KA);
 })
 
mod <- RxODE({
     a = 6
     b = 0.6
     d/dt(intestine) = -a*intestine
     d/dt(blood)     = a*intestine - b*blood
})
et1 <- eventTable(time.units="days")
et1$add.sampling(seq(0,10,by=1/24))
et1$add.dosing(dose=2/24,rate=2,strt.time=0,
               nbr.doses=10,dosing.interval=1)
 


```


## Benchmarking the Models
These benchmark the single solve
```{r}

library(microbenchmark)

res <- microbenchmark(solved.1c=rxSolve(solve.1c, params=c(V=20, CL=25), events=et), 
                      ode.1c.lsoda=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="lsoda"),
                      ode.1c.liblsoda=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="liblsoda"),
                      ode.1c.dop=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="dop853"),
					  ##
                      solved.1c.ka=rxSolve(solve.1c.ka, params=c(V=20, CL=25, KA=2), events=et), 
                      ode.1c.ka.lsoda=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="lsoda"),
                      ode.1c.ka.liblsoda=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="liblsoda"),
                      ode.1c.ka.dop=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="dop853"),
					  ##
					  solved.2c=rxSolve(solve.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et), 
                      ode.2c.lsoda=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="lsoda"),
                      ode.2c.liblsoda=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="liblsoda"),
                      ode.2c.dop=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="dop853"),
					  ##
					  solved.2c.ka=rxSolve(solve.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et), 
                      ode.2c.ka.lsoda=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="lsoda"),
                      ode.2c.ka.liblsoda=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="liblsoda"),
                      ode.2c.ka.dop=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="dop853"),
					  ## 
					  solved.3c=rxSolve(solve.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et), 
                      ode.3c.lsoda=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="lsoda"),
                      ode.3c.liblsoda=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="liblsoda"),
                      ode.3c.dop=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="dop853"),
					  ##
					  solved.3c.ka=rxSolve(solve.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et), 
                      ode.3c.ka.lsoda=rxSolve(ode.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="lsoda"),
                      ode.3c.ka.liblsoda=rxSolve(ode.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="liblsoda"),
                      ode.3c.ka.dop=rxSolve(ode.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="dop853"),
					  data.par.setup=rxDataParSetup(mod, et)
					  )

```

0.6 solve speeds

Unit: microseconds
               expr      min        lq      mean    median        uq        max
          solved.1c  738.986  827.3425 2113.2713  906.5515 1271.5820 104919.876
       ode.1c.lsoda  478.378  545.0920  762.3067  608.4585  918.1540   4823.488
    ode.1c.liblsoda  482.839  550.4460 1812.7325  623.6305  914.5845 109069.976
         ode.1c.dop  477.039  543.0835  750.8649  646.8355  933.7725   1460.568
       solved.1c.ka  917.931  999.1480 2705.8966 1084.6040 1593.1035 136515.969
    ode.1c.ka.lsoda  525.233  589.7160  778.9606  665.1320  949.3915   1498.498
 ode.1c.ka.liblsoda  535.943  614.0365  842.6000  747.0185 1052.2515   2097.363
      ode.1c.ka.dop  526.126  607.1200 2018.6935  703.5090  970.5880 119313.582
          solved.2c 1112.941 1241.0140 2861.2263 1362.1705 1945.8620 131661.245
       ode.2c.lsoda  546.207  616.2680  851.7034  713.9955 1009.8575   5453.142
    ode.2c.liblsoda  541.744  620.0610 1872.2843  723.3670 1026.1460 103915.820
         ode.2c.dop  542.637  609.3510 2003.3560  711.5415  953.1840 118396.098
       solved.2c.ka 1269.574 1425.7600 3159.7523 1557.1800 2200.6685 126936.379
    ode.2c.ka.lsoda  580.122  649.9595  933.2237  774.4625 1108.0325   4628.923
 ode.2c.ka.liblsoda  572.089  670.7095 2320.7093  822.6570 1095.7605 138860.553
      ode.2c.ka.dop  582.799  633.2250  836.6693  695.2535  993.5700   4266.571
          solved.3c 1535.984 1699.7560 3176.5000 1873.7925 2632.6360 107678.131
       ode.3c.lsoda  581.014  673.3880  844.2689  751.0345 1025.7000   1555.172
    ode.3c.liblsoda  602.435  668.4785 2305.2870  722.6980 1097.3225 136529.803
         ode.3c.dop  595.294  664.2395  863.1273  740.9940 1000.4870   2966.206
       solved.3c.ka 1712.697 1869.1075 3690.9294 2067.9105 2939.6540 130059.664
    ode.3c.ka.lsoda  593.956  665.5785  947.1735  771.3385 1070.9940   6849.450
 ode.3c.ka.liblsoda  597.079  682.0890  927.2531  794.3205 1116.9570   4032.737
      ode.3c.ka.dop  601.988  687.8905 2239.6395  798.7835 1145.5170 132009.318


