---
title: "Nesting levels in RxODE"
output:
  html_document:
    self_contained: no
date: "`r Sys.Date()`"
---

```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
knitr::opts_chunk$set( comment = "#>")
options(width=80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH=80)
```
## Nesting in RxODE

More than one level of nesting is possible in RxODE; In this example
we will be using the following uncertainties and sources of variability:

|                     Level | Variable           | Matrix specified | Integrated Matrix |
|--------------------------:|--------------------|------------------|-------------------|
|         Model uncertainty | NA                 | `thetaMat`       | `thetaMat`        |
|              Investigator | `inv.Cl`, `inv.Ka` | `omega`          | `theta`           |
|                   Subject | `eta.Cl`, `eta.Ka` | `omega`          | `omega`           |
|                       Eye | `eye.Cl`, `eye.Ka` | `omega`          | `omega`           |
|                  Occasion | `iov.Cl`, `occ.Ka` | `omega`          | `omega`           |
| Unexplained Concentration | `prop.sd`          | `sigma`          | `sigma`           |
|        Unexplained Effect | `add.sd`           | `sigma`          | `sigma`           |


### Event table

This event table contains nesting variables:
 - inv: investigator id
 - id: subject id
 - eye: eye id (left or right)
 - occ: occasion
 
```{r}
library(dplyr)
et(amountUnits="mg", timeUnits="hours") %>%
  et(amt=10000, addl=9,ii=12,cmt="depot") %>%
  et(time=120, amt=2000, addl=4, ii=14, cmt="depot") %>%
  et(seq(0, 240, by=4)) %>% # Assumes sampling when there is no dosing information
  et(seq(0, 240, by=4) + 0.1) %>% ## adds 0.1 for separate eye
  et(id=1:20) %>%
  ## Add an occasion per dose
  mutate(occ=cumsum(!is.na(amt))) %>%
  mutate(occ=ifelse(occ == 0, 1, occ)) %>%
  mutate(occ=2- occ %% 2) %>%
  mutate(eye=ifelse(round(time) == time, 1, 2)) %>%
  mutate(inv=ifelse(id < 10, 1, 2)) %>% as.tbl ->
  ev
```

### RxODE model

This creates the `RxODE` model with multi-level nesting.  Note the
variables `inv.Cl`, `inv.Ka`, `eta.Cl` etc; You only need one variable
for each level of nesting.

```{r}
mod <- RxODE({
  ## Clearance with individuals
  eff(0) = 1
  C2 = centr/V2*(1+prop.sd);
  C3 = peri/V3;
  CL =  TCl*exp(eta.Cl + eye.Cl + iov.Cl + inv.Cl)
  KA = TKA * exp(eta.Ka + eye.Ka + iov.Cl + inv.Ka)
  d/dt(depot) =-KA*depot;
  d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
  d/dt(peri)  =                    Q*C2 - Q*C3;
  d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
  ef0 = eff + add.sd
})
```

### Uncertainty in Model parameters

```{r}
theta <- c("TKA"=0.294, "TCl"=18.6, "V2"=40.2,
           "Q"=10.5, "V3"=297, "Kin"=1, "Kout"=1, "EC50"=200)

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(names(theta), names(theta))

tMat
```

### Nesting Variability 

To specify multiple levels of nesting, you can specify it as a nested
`lotri` matrix; When using this approach you use the condition
operator `|` to specify what variable nesting occurs on; For the
Bayesian simulation we need to specify how much information we have
for each parameter; For `RxODE` this is the `nu` parameter.

In this case:
- id, `nu=100` or the model came from 100 subjects
- eye, `nu=200` or the model came from 200 eyes
- occ, `nu=200` or the model came from 200 occasions
- inv, `nu=10` or the model came from 10 investigators

To specify this in `lotri` you can use `| var(nu=X)`, or:

```{r}
omega <- lotri(lotri(eta.Cl ~ 0.1,
                     eta.Ka ~ 0.1) | id(nu=100),
               lotri(eye.Cl ~ 0.05,
                     eye.Ka ~ 0.05) | eye(nu=200),
               lotri(iov.Cl ~ 0.01,
                     iov.Ka ~ 0.01) | occ(nu=200),
               lotri(inv.Cl ~ 0.02,
                     inv.Ka ~ 0.02) | inv(nu=10))
omega
```

### Unexplained variability
The last piece of variability to specify is the unexplained variability
```{r}
sigma <- lotri(prop.sd ~ .25,
               add.sd~ 0.125)
```
### Solving the problem

```{r}
s <- rxSolve(mod, theta, ev,
             thetaMat=tMat, omega=omega,
             sigma=sigma, sigmaDf=400,
             nStud=400)
```
