---
title: "RxODE Simulation"
author: "Matthew L. Fidler, Melissa Hallow, and Wenping Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_width: 850px
fig_height: 500px
vignette: >
  %\VignetteIndexEntry{RxODE Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
```


# Simulation of Variability with RxODE
Variability in model parameters can be simulated by creating a matrix
of parameter values for use in the simulation. In the example below,
40% variability in clearance is simulated.

```{r}
set.seed(42);
library(RxODE)
mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2;
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
})

theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
               Q=1.05E+01, V3=2.97E+02,                # peripheral
               Kin=1, Kout=1, EC50=200)                # effects  
```

Each subproblem can be simulated by using the rxSolve function to run
the simulation for each set of parameters of in the parameter matrix.

```{r fig.width=10}
## the column names of the omega matrix need to match the parameters specified by RxODE
omega <- matrix(0.4^2,dimnames=list(NULL,c("eta.Cl")))

ev <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=100));

sim  <- rxSolve(mod,theta,ev,omega=omega,nSub=100)

library(ggplot2)
library(gridExtra)

p1 <- ggplot(sim,aes(time,centr,color=factor(sim.id))) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)") + guides(color=FALSE)

p2 <-ggplot(sim,aes(time,eff,color=factor(sim.id))) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

It is now straightforward to perform calculations and generate plots
with the simulated data. Below,  the 5th, 50th, and 95th percentiles
of the simulated data are plotted. 

```{r fig.width=10}

library(dplyr)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$eff, probs=p), 
                  eff.n = length(.$eff), eff.avg = mean(.$eff),
                  centr=quantile(.$centr, probs=p),
                  centr.n=length(.$centr),centr.avg = mean(.$centr))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```


Note that you can see the parameters that were simulated for the example

```{r}
head(sim$param)
```

You can also supply a data-frame of parameters to simulate instead of
using an omega simulation.  In this contrived example we will use the
previously simulated data.

```{r,results="asis"}
theta <- sim$param;
sim  <- rxSolve(mod,theta,ev)
rxHtml(sim)
```

Even though multiple subjects were simulated, this is still a reactive
data frame, meaning you can change things about the model on the fly.

For example, if the effect at time 0 should have been 100, you can fix
this by:

```{r,results="asis"}
sim$eff0 <- 100
rxHtml(sim)
```

#### Simulation of unexplained variability 

In addition to conveniently simulating between subject variability,
you can also easily simulate unexplained variability.

```{r}
mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2;
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
    e = eff+eff.err
    cp = centr*(1+cp.err)
})

theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
           Q=1.05E+01, V3=2.97E+02,                # peripheral
           Kin=1, Kout=1, EC50=200)                # effects  

sigma <- diag(2)*0.1
dimnames(sigma) <- list(NULL, c("eff.err","cp.err"))


sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```
#### Simulation of Individuals

Sometimes you may want to match the dosing and observations of
individuals in a clinical trial.  To do this you will have to create a
data.frame using the `RxODE` event specification as well as an `ID`
column to indicate an individual. The RxODE event vignette talks more about
how these datasets should be created.

If you have a NONMEM/Monlix dataset and the package
`nlmixr`, you can convert the `NONMEM`` dataset to a `RxODE` compatible dataset to
use for simulation with the `nmDataConvert` function.

Instead of using nlmixr for this simple example, I will combine two
RxODE event tables.

```{r}

ev1 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=10));

ev2 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=5000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=8));

dat <- rbind(data.frame(ID=1, ev1$get.EventTable()),
             data.frame(ID=2, ev2$get.EventTable()))


## Note the number of subject is not needed since it is determined by the data
sim  <- rxSolve(mod, theta, dat, omega=omega, sigma=sigma)

sim %>% select(id, time, e, cp)
```

#### Simulation of Clinical Trials

By either using a simple single event table, or data from a clinical
trial as described above, a complete clinical trial simulation can be
performed.

Typically in clinical trial simulations you want to account for the
uncertainty in the fixed parameter estimates, and even the uncertainty
in both your between subject variability as well as the unexplained
variability.

`RxODE` allows you to account for these uncertainties by simulating
multiple virtual "studies," specified by the parameter `nStud`.  In a
single virtual study:

- A Population effect parameter is sampled from a multivariate normal
  distribution with mean given by the parameter estimates and the
  variance specified by the named matrix `thetaMat`.
  
- A between subject variability/covariance matrix is sampled from
  either a scaled inverse chi-squared distribution (for the univariate
  case) or a inverse Wishart that is parameterized to scale to the
  conjugate prior covariance term, as described by
  the
  [wikipedia article](https://en.wikipedia.org/wiki/Scaled_inverse_chi-squared_distribution). (This
  is not the same as
  the
  [scaled inverse Wishart distribution](http://andrewgelman.com/2012/08/22/the-scaled-inverse-wishart-prior-distribution-for-a-covariance-matrix-in-a-hierarchical-model/) ).
  In the case of the between subject variability, the
  variance/covariance matrix is given by the 'omega' matrix parameter
  and the degrees of freedom is the number of subjects in the
  simulation.
  
- Unexplained variability is also simulated from the scaled inverse
  chi squared distribution or inverse Wishart distribution with the
  variance/covariance matrix given by the 'sigma' matrix parameter and
  the degrees of freedom given by the number of observations being
  simulated.
  
The covariance/variance prior is simulated from `RxODE`s `cvPost` function.

An example of this simulation is below:

```{r}

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(NULL, names(theta))

sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
     dfSub=10, dfObs=100)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)

```
If you wish you can see what `omega` and `sigma` was used for each
virtual study by accessing them in the solved data object with
`$omega.list` and `$sigma.list`:

```{r}
head(sim$omega.list)
```

```{r}
head(sim$sigma.list)
```

You can also see the parameter realizations from the `$params` data frame.

If you do not wish to sample from the prior distributions of either
the `omega` or `sigma` matrices, you can turn off this feature by
specifying the `simVariability = FALSE` option when solving:

```{r}
sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                simVariability=FALSE);

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))


p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

Note since realizations of `omega` and `sigma` were not simulated, `$omega.list` and
`$sigma.list` both return `NULL`.

#### RxODE multi-threaded solving and simulation

RxODE now supports multi-threaded solving on OpenMP supported
compilers, including linux and windows. Mac OSX can also be supported
but
takes
[additional setup](https://github.com/Rdatatable/data.table/wiki/Installation#openmp-enabled-compiler-for-mac).
By default it uses all your available cores for solving as determined
by `rxCores()`.  This may be overkill depending on your system, at a
certain point the speed of solving is limited by things other than
computing power.

You can also speed up simulation by using the multi-cores to generate
random deviates with `mvnfast`.  This is controlled by the `nCoresRV`
parameter.  For example:

```{r}
sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                nCoresRV=2);

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))


p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

The default for this is `1` core since the result depends on the
number of cores and the random seed you use in your simulation.
However, you can always speed up this process with more cores if you
are sure your collaborators have the same number of cores available to
them and have OpenMP thread-capable compile.

