---
title: "RxODE Benchmarking"
author: "Matthew L. Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RxODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Setup the RxODE models and event table for comparison

```{r}
library(RxODE)

ver <- packageVersion("RxODE")

ver.no <- ver$minor

et <- eventTable();
et$add.dosing(dose=3, nbr.doses=6, dosing.interval=8)
et$add.sampling(seq(0, 48, length.out=200))

if (ver.no > 5){
    solve.1c <- RxODE({
        C2 = linCmt(CL, V);
    })
    solve.1c.ka <- RxODE({
        C2 = linCmt(V, CL, KA);
    })
    solve.2c <- RxODE({
        C2=linCmt(V, CL, V2, Q);
    })
    solve.2c.ka <- RxODE({
        C2=linCmt(V, CL, V2, Q, KA);
    })
    solve.3c <- RxODE({
        ## double solvedC(double t, int parameterization, int cmt, unsigned int col, double p1, double p2, double p3, double p4, double p5, double p6, double p7, double p8);
        C2=linCmt(V, CL, V2, Q, Q2, V3);
    })
    solve.3c.ka <- RxODE({
        ## double solvedC(double t, int parameterization, int cmt, unsigned int col, double p1, double p2, double p3, double p4, double p5, double p6, double p7, double p8);
        C2=linCmt(V, CL, V2, Q, Q2, V3, KA);
    })
}

ode.1c <- RxODE("
C2 = center/V;
d/dt(center) = - CL*C2;
","ode1c")

ode.1c.ka <- RxODE("
    C2 = center/V;
    d/dt(depot) = -KA * depot;
    d/dt(center) = KA * depot - CL*C2;
","ode1cka")

ode.2c <- RxODE("
    C2 = centr/V;
    C3 = peri/V2;
    d/dt(centr) = - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  = Q*C2 - Q*C3;
", "ode2c")


ode.2c.ka <- RxODE("
   C2 = centr/V;
   C3 = peri/V2;
   d/dt(depot) =-KA*depot;
   d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
   d/dt(peri)  =                    Q*C2 - Q*C3;
","ode2cka")

ode.3c <- RxODE("
     C2 = centr/V;
     C3 = peri/V2;
     C4 = peri2 / V3;
     d/dt(centr) = - CL*C2 - Q*C2 + Q*C3  - Q2*C2 + Q2*C4;
     d/dt(peri)  = Q*C2 - Q*C3;
     d/dt(peri2) = Q2 * C2 - Q2 * C4;
 ","ode3c")
 
 ode.3c.ka <- RxODE("
     C2 = centr/V;
     C3 = peri/V2;
     C4 = peri2 / V3;
	 d/dt(depot) =-KA*depot;
     d/dt(centr) = KA*depot- CL*C2 - Q*C2 + Q*C3  - Q2*C2 + Q2*C4;
     d/dt(peri)  = Q*C2 - Q*C3;
     d/dt(peri2) = Q2 * C2 - Q2 * C4;
 ","ode3cka")

```


## Benchmarking the Models
These benchmark the single solve
```{r}

library(microbenchmark)

if (ver.no > 5){
    res.rxSolve <- microbenchmark(solved.1c=rxSolve(solve.1c, params=c(V=20, CL=25), events=et),
                                  solved.1c.ka=rxSolve(solve.1c.ka, params=c(V=20, CL=25, KA=2), events=et),
                                  solved.2c=rxSolve(solve.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et),
                                  solved.2c.ka=rxSolve(solve.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et),
                                  solved.3c=rxSolve(solve.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et),
                                  solved.3c.ka=rxSolve(solve.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et) 
                                  )
}

if (ver.no < 7){
    res.run <- microbenchmark(
        ode.1c.lsoda=ode.1c$run(params=c(V=20, CL=25), events=et, stiff=TRUE),
        ode.1c.dop=ode.1c$run(params=c(V=20, CL=25), events=et, stiff=FALSE),
        ode.1c.ka.lsoda=ode.1c.ka$run(params=c(V=20, CL=25, KA=2), events=et, stiff=TRUE),
        ode.1c.ka.dop=ode.1c.ka$run(params=c(V=20, CL=25, KA=2), events=et, stiff=FALSE),
        ode.2c.lsoda=ode.2c$run(params=c(V=20, CL=25, V2=297, Q=10), events=et, stiff=TRUE),
        ode.2c.dop=ode.2c$run(params=c(V=20, CL=25, V2=297, Q=10), events=et, stiff=FALSE),
        ode.2c.ka.lsoda=ode.2c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, stiff=TRUE),
        ode.2c.ka.dop=ode.2c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, stiff=FALSE),
        ode.3c.lsoda=ode.3c$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, stiff=TRUE),
        ode.3c.dop=ode.3c$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, stiff=FALSE),
        ode.3c.ka.lsoda=ode.3c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, stiff=TRUE),
        ode.3c.ka.dop=ode.3c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, stiff=FALSE))
} else {
    res.run <- microbenchmark(
        ode.1c.lsoda=ode.1c$run(params=c(V=20, CL=25), events=et, method="lsoda"),
        ode.1c.liblsoda=ode.1c$run(params=c(V=20, CL=25), events=et, method="liblsoda"),
        ode.1c.dop=ode.1c$run(params=c(V=20, CL=25), events=et, method="dop853"),
        ode.1c.ka.lsoda=ode.1c.ka$run(params=c(V=20, CL=25, KA=2), events=et, method="lsoda"),
        ode.1c.ka.liblsoda=ode.1c.ka$run(params=c(V=20, CL=25, KA=2), events=et, method="liblsoda"),
        ode.1c.ka.dop=ode.1c.ka$run(params=c(V=20, CL=25, KA=2), events=et, method="dop853"),
        ode.2c.lsoda=ode.2c$run(params=c(V=20, CL=25, V2=297, Q=10), events=et, method="lsoda"),
        ode.2c.liblsoda=ode.2c$run(params=c(V=20, CL=25, V2=297, Q=10), events=et, method="liblsoda"),
        ode.2c.dop=ode.2c$run(params=c(V=20, CL=25, V2=297, Q=10), events=et, method="dop853"),
        ode.2c.ka.lsoda=ode.2c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="lsoda"),
        ode.2c.ka.liblsoda=ode.2c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="liblsoda"),
        ode.2c.ka.dop=ode.2c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="dop853"),
        ode.3c.lsoda=ode.3c$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="lsoda"),
        ode.3c.liblsoda=ode.3c$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="liblsoda"),
        ode.3c.dop=ode.3c$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="dop853"),
        ode.3c.ka.lsoda=ode.3c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="lsoda"),
        ode.3c.ka.liblsoda=ode.3c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="liblsoda"),
        ode.3c.ka.dop=ode.3c.ka$run(params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="dop853"))
    res.run.df <- microbenchmark(
        ode.1c.lsoda=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="lsoda", return.type="data.frame"),
        ode.1c.liblsoda=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="liblsoda", return.type="data.frame"),
        ode.1c.dop=rxSolve(ode.1c, params=c(V=20, CL=25), events=et, method="dop853", return.type="data.frame"),
        ode.1c.ka.lsoda=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="lsoda", return.type="data.frame"),
        ode.1c.ka.liblsoda=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="liblsoda", return.type="data.frame"),
        ode.1c.ka.dop=rxSolve(ode.1c.ka, params=c(V=20, CL=25, KA=2), events=et, method="dop853", return.type="data.frame"),
        ode.2c.lsoda=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="lsoda", return.type="data.frame"),
        ode.2c.liblsoda=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="liblsoda", return.type="data.frame"),
        ode.2c.dop=rxSolve(ode.2c, params=c(V=20, CL=25, V2=297, Q=10), events=et, method="dop853", return.type="data.frame"),
        ode.2c.ka.lsoda=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="lsoda", return.type="data.frame"),
        ode.2c.ka.liblsoda=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="liblsoda", return.type="data.frame"),
        ode.2c.ka.dop=rxSolve(ode.2c.ka, params=c(V=20, CL=25, V2=297, Q=10, KA=2), events=et, method="dop853", return.type="data.frame"),
        ode.3c.lsoda=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="lsoda", return.type="data.frame"),
        ode.3c.liblsoda=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="liblsoda", return.type="data.frame"),
        ode.3c.dop=rxSolve(ode.3c, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400), events=et, method="dop853", return.type="data.frame"),
        ode.3c.ka.lsoda=rxSolve(ode.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="lsoda", return.type="data.frame"),
        ode.3c.ka.liblsoda=rxSolve(ode.3c.ka, params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="liblsoda", return.type="data.frame"),
        ode.3c.ka.dop=rxSolve(ode.3c.ka,params=c(V=20, CL=25, V2=297, Q=10, Q2=7, V3=400, KA=2), events=et, method="dop853", return.type="data.frame"))
}


```

### 0.5 Solve speeds
0.5.3 solve speeds

(These should be run on the same machine, and were run on my machine;  CRAN machines may give different results)

Unit: microseconds
            expr     min       lq     mean   median        uq      max neval
    ode.1c.lsoda 347.627 390.9135 551.1030 438.6620  715.3360 1004.950   100
      ode.1c.dop 511.401 595.7410 834.2383 866.6135 1006.0660 1671.645   100
 ode.1c.ka.lsoda 360.568 409.4325 612.0649 527.9115  770.4475 1848.805   100
   ode.1c.ka.dop 517.201 601.9885 940.1150 907.4450 1092.6385 4294.690   100
    ode.2c.lsoda 380.203 425.2745 606.0941 541.9680  791.6435  965.234   100
      ode.2c.dop 525.680 578.7840 865.9800 709.3110 1047.5675 3695.825   100
 ode.2c.ka.lsoda 402.515 464.5445 709.5700 730.5075  846.5330 2250.428   100
   ode.2c.ka.dop 535.051 618.2770 929.9272 995.1330 1110.2645 3730.632   100
    ode.3c.lsoda 419.919 505.3765 769.5949 781.1570  918.6020 2007.223   100
      ode.3c.dop 553.348 627.6475 920.1455 799.4530 1068.7640 4702.114   100
 ode.3c.ka.lsoda 418.135 479.9395 677.4090 539.9600  871.2995 1700.651   100
   ode.3c.ka.dop 576.998 662.4550 953.7390 911.4620 1166.7150 3182.640   100

### 0.6 solve speeds

   
### 0.7 solve speeds

(rxSolve object)

> res.run
Unit: microseconds
               expr     min       lq      mean   median        uq       max
       ode.1c.lsoda 502.922 537.2825  715.7685 591.9480  909.4535  1816.675
    ode.1c.liblsoda 521.664 552.6785  744.0963 618.0535  841.1775  3079.110
         ode.1c.dop 401.178 444.9095  567.4981 477.7090  629.8795  1828.278
    ode.1c.ka.lsoda 531.035 581.4615  764.2354 629.8795  962.1105  1356.593
 ode.1c.ka.liblsoda 590.833 630.3250  830.7398 765.3155  963.6730  2482.477
      ode.1c.ka.dop 454.727 494.6660  629.8211 543.9760  754.3820  1300.367
       ode.2c.lsoda 537.729 589.2710  764.1283 625.1935  991.7860  1331.158
    ode.2c.liblsoda 601.096 630.1020 1287.7422 689.0070  917.9320 44045.559
         ode.2c.dop 461.867 512.2925  625.1265 540.8530  714.8890  1130.792
    ode.2c.ka.lsoda 564.504 626.7555  775.5078 663.1250  877.3235  1595.783
 ode.2c.ka.liblsoda 643.490 676.5125  892.7102 734.0780  953.4090  4341.099
      ode.2c.ka.dop 495.335 529.6965  669.7514 572.5355  730.9550  1216.026
       ode.3c.lsoda 579.230 626.7550  758.1976 674.9505  792.3135  1771.604
    ode.3c.liblsoda 645.275 683.2055  920.6498 768.6620 1045.7820  4960.938
         ode.3c.dop 508.723 557.8100  697.0037 614.7070  798.3375  1271.806
    ode.3c.ka.lsoda 606.898 659.5545  937.0584 726.2685 1098.6625  7517.046
 ode.3c.ka.liblsoda 668.926 702.8405  950.7582 761.2995 1027.2630  5412.987
      ode.3c.ka.dop 529.696 583.9160  762.1783 676.7350  956.5325  1344.098

Data frame object
> res.run.df
Unit: microseconds
               expr     min       lq     mean   median       uq       max neval
       ode.1c.lsoda 479.270 515.6400 659.1886 549.7770 810.6090  1253.956   100
    ode.1c.liblsoda 486.410 541.5215 714.6930 606.6740 872.4145  1431.563   100
         ode.1c.dop 373.510 412.5570 531.8384 442.2315 619.8385  1844.789   100
    ode.1c.ka.lsoda 505.599 558.7030 755.0517 646.8370 898.2975  1957.690   100
 ode.1c.ka.liblsoda 560.488 597.0800 737.9381 650.4065 814.8490  1258.419   100
      ode.1c.ka.dop 422.151 466.1065 604.0682 517.6480 737.2020  1328.927   100
       ode.2c.lsoda 511.400 554.4630 723.1092 631.4410 837.8305  2586.899   100
    ode.2c.liblsoda 579.230 607.5670 769.5324 685.2135 889.8185  1670.306   100
         ode.2c.dop 435.984 486.8570 648.8581 560.9340 790.7515  1608.278   100
    ode.2c.ka.lsoda 531.481 568.0735 736.1441 617.8305 947.6075  1550.711   100
 ode.2c.ka.liblsoda 608.237 640.3660 817.8566 673.1655 914.3620  3078.218   100
      ode.2c.ka.dop 468.114 513.8545 729.6423 653.0840 900.9750  2146.453   100
       ode.3c.lsoda 543.530 584.8080 785.2538 625.4165 922.6175  3519.558   100
    ode.3c.liblsoda 633.226 662.4545 836.5320 701.9475 863.7130  4795.827   100
         ode.3c.dop 483.287 520.3250 699.7169 625.4165 836.4915  1783.206   100
    ode.3c.ka.lsoda 581.461 620.2850 774.5928 677.1815 828.2360  1986.250   100
 ode.3c.ka.liblsoda 655.538 687.4450 981.1119 743.4490 992.2325 12715.388   100
      ode.3c.ka.dop 497.120 545.3155 685.4102 584.5855 804.3615  1288.764   100

## Backward old-c interface
This is a measure of how fast the SAEM algorithm will run in nlmixr
```
demo <- structure(list(ID = c(1012, 1012, 1012, 1012, 1012, 1012, 1012, 1012), TIME = c(588.5, 600.5, 612.5, 624.5, 636.5, 648.5, 660.5, 672.5), DOSE = c(2000, 2000, 2000, 2000, 2000, 2000, 2000, 2000), AMT = c(1000, 1000, 1000, 1000, 1000, 1000, 1000, 0), TAD = c(0, 0, 0, 0, 0, 0, 0, 12), CL = c(5.851496056, 5.851496056, 5.851496056, 5.851496056, 5.851496056, 5.851496056, 5.851496056, 5.851496056), V = c(49.3930186, 49.3930186, 49.3930186, 49.3930186, 49.3930186, 49.3930186, 49.3930186, 49.3930186), KA = c(3.320205555, 3.320205555, 3.320205555, 3.320205555, 3.320205555, 3.320205555, 3.320205555, 3.320205555), TAD4 = c(0, 12, 24, 36, 48, 60, 72, 84)), .Names = c("ID", "TIME", "DOSE", "AMT", "TAD", "CL", "V", "KA", "TAD4"), row.names = c(NA, -8L), class = c("data.frame"), sorted = c("ID", "TIME"));

ode1KA <- "
d/dt(abs)    = -KA*abs;
d/dt(centr)  =  KA*abs-(CL/V)*centr;
C1=centr/V;
"

StepSize=1
Extension=6

mod1KA <- RxODE(model=ode1KA)

params <- demo[1,c("CL","V","KA")]

ev<-eventTable()
DOSi<-as.data.frame(demo[demo$AMT>0,])
for (j in 1:length(DOSi$AMT)){
    dos<-DOSi[j,]
    ev$add.dosing(dose=as.numeric(dos$AMT),nbr.doses=1,dosing.to=1,rate=NULL,start.time=as.numeric(dos$TAD4))
}
timei<-demo$TIME
minimum<-min(timei)
maximum<-max(timei)+Extension
times<-sort(unique(c(timei,seq(minimum,maximum,StepSize))))
ev$add.sampling(times)

x <- as.data.frame(mod1KA$run(params, ev))

## test old solving.
event.table <- ev$get.EventTable()
modelVars <- mod1KA$get.modelVars()
state_vars <- modelVars$state;
neq <- length(state_vars);
lhs_vars <- modelVars$lhs;
nlhs <- length(lhs_vars);

ntime <- dim(event.table)[1];
ret <- rep(0.0, ntime * neq);
lhs <- rep(0.0, ntime * nlhs);
rc <- as.integer(0)

inits <- rep(0.0, neq)

cmpMgr <- mod1KA$cmpMgr;

## cmpMgr$dynLoad()

atol <- 1e-8
rtol <- 1e-6
transit_abs <- FALSE
stiff <- TRUE

ode_solve <- cmpMgr$ode_solver;

library(microbenchmark)

res <- microbenchmark(rxode.lsoda = mod1KA$dll$.c(ode_solve,
                                                  as.integer(neq),
                                                  as.double(params),
                                                  as.double(event.table$time),
                                                  as.integer(event.table$evid),
                                                  length(event.table$time),
                                                  as.double(inits),
                                                  as.double(event.table$amt[event.table$evid > 0]),
                                                  as.double(ret),
                                                  as.double(atol),
                                                  as.double(rtol),
                                                  as.integer(TRUE),
                                                  as.integer(transit_abs),
                                                  as.integer(nlhs),
                                                  as.double(lhs),
                                                  rc),
                      rxode.dop = mod1KA$dll$.c(ode_solve,
                                                  as.integer(neq),
                                                  as.double(params),
                                                  as.double(event.table$time),
                                                  as.integer(event.table$evid),
                                                  length(event.table$time),
                                                  as.double(inits),
                                                  as.double(event.table$amt[event.table$evid > 0]),
                                                  as.double(ret),
                                                  as.double(atol),
                                                  as.double(rtol),
                                                  as.integer(FALSE),
                                                  as.integer(transit_abs),
                                                  as.integer(nlhs),
                                                  as.double(lhs),
                                                rc),
                      times=5000L
                      )
```
### 0.6 speed
> res
Unit: microseconds
        expr      min       lq      mean   median       uq       max neval
 rxode.lsoda  465.883  520.102  683.5532  585.032  838.723  9688.488  5000
   rxode.dop 5453.596 5865.928 6381.3596 6036.396 6751.062 44734.565  5000

### 0.7 speed

Unit: microseconds
        expr     min       lq     mean   median       uq      max neval
 rxode.lsoda  447.14  496.674  632.594  548.439  766.431  4618.22  5000
   rxode.dop 5288.93 5667.348 6113.333 5797.206 6430.432 23199.09  5000
 
