---
title: "RxODE Events"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RxODE Events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
knitr::opts_chunk$set( comment = "#>")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
options(width=80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH=80)
```

## RxODE event tables
In general, RxODE event tables follow NONMEM convention with the exceptions:

 - The compartment data item (`cmt`) can be a string/factor with compartment names
   - You may turn off a compartment with a negative compartment number
     or "-cmt" where cmt is the compartment name.
   - The compartment data item (`cmt`) can still be a number, the
     number of the compartment is defined by the appearance of the
     compartment name in the model. This can be tedious to count, so
     you can specify compartment numbers easier by using the
     `cmt(cmtName)` at the beginning of the model.
 - An additional column, `dur` can specify the duration of infusions; 
    - Bioavailability changes will change the rate of infusion since
      `dur`/`amt` are fixed in the input data.
    - Similarly, when specifying `rate`/`amt` for an infusion, the
      bioavailability will change the infusion duration since
      `rate`/`amt` are fixed in the input data.
 - Some infrequent NONMEM columns are not supported: `pcmt`, `call`.
 
 Here are the legal entries to a data table:
 
| Data Item | Meaning               | Notes                                                              |
|-----------|-----------------------|--------------------------------------------------------------------|
| id        | Individual identifier | Must be an integer > 0, sorted                                     |
| time      | Individual time       | For each ID must be ascending and non-negative                     |
| amt       | dose amount           | Positive for doses zero/NA for observations                        |
| rate      | infusion rate         | When specified the infusion duration will be dur=amt/rate          |
|           |                       | rate = -1, rate modeled; rate = -2, duration modeled               |
| dur       | infusion duration     | When specified the infusion rate will be rate = amt/dur            |
| evid      | event ID              | 0=Observation; 1=Dose; 2=Other; 3=Reset; 4=Reset+Dose              |
| cmt       | Compartment           | Represents compartment #/name for dose/observation                 |
| ss        | Steady State Flag     | 0 = non-steady-state; 1=steady state; 2=steady state +prior states |
| ii        | Inter-dose Interval   | Time between doses.                                                |
| addl      | # of additional doses | Number of doses like the current dose.                             |


Other notes:

 - The `evid` can be the classic RxODE (described at the end of this
   document) or the NONMEM-style evid described above.
 - NONMEM's `DV` is not required; RxODE is a ODE solving framework.
 - NONMEM's `MDV` is not required, since it is captured in `EVID`
 
When returning the RxODE solved data-set there are a few additional
event ids (EVID):

- EVID = -1 is when a modeled rate ends (corresponds to rate = -1)
- EVID = -2 is when a modeled duration ends (corresponds to rate=-2)
- EVID = -10 when a rate specified zero-order infusion ends (corresponds to rate > 0)
- EVID = -20 when a duration specified zero-order infusion ends (corresponds to dur > 0)
- EVID = 101, 102, 103,... These correspond to the 1, 2, 3, ... modeled time (mtime).

These can only be accessed when solving with the option combination
`addDosing=TRUE` and `subsetNonmem=FALSE`. If you want to see the
classic `EVID` equivalents you can use `addDosing=NA`.
 
## Creating RxODE's event tables
An event table in RxODE is a specialized data frame that acts as a
container for all of RxODE's events and observation times.

To create an RxODE event table you may use the code `eventTable()`,
`et()`, or even create your own data frame with the right event
information contained in it.

```{r}
library(RxODE)
(ev <- eventTable());
```

or

```{r}
(ev <- et());
```

For these models, we can illustrate by using the model shared in the RxODE tutorial:

```{r}
## Model from RxODE tutorial
m1 <-RxODE({
    KA=2.94E-01;
    CL=1.86E+01;
    V2=4.02E+01;
    Q=1.05E+01;
    V3=2.97E+02;
    Kin=1;
    Kout=1;
    EC50=200;
    ## Added modeled bioavaiblity, duration and rate
    fdepot = 1;
    durDepot = 8;
    rateDepot = 1250;
    C2 = centr/V2;
    C3 = peri/V3;
    d/dt(depot) =-KA*depot;
    f(depot) = fdepot
    dur(depot) = durDepot
    rate(depot) = rateDepot
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
    eff(0) = 1
});
```

### Adding doses to the event table

Once created you can add dosing to the event table by the
`add.dosing`, and `et` functions.

Using the `add.dosing` function you have:

| argument        | meaning                                 |
|-----------------|-----------------------------------------|
| dose            | dose amount                             |
| nbr.doses       | Number of doses;  Should be at least 1. |
| dosing.interval | Dosing interval; By default this is 24. |
| dosing.to       | Compartment where dose is administered. |
| rate            | Infusion rate                           |
| start.time      | The start time of the dose              |

```{r}
ev <- eventTable(amount.units="mg", time.units="hr")

## The methods ar attached to the event table, so you can use them
## directly
ev$add.dosing(dose=10000, nbr.doses = 3)# loading doses
## Starts at time 0; Default dosing interval is 24

## You can also pipe the event tables to these methods.
ev <- ev %>%
    add.dosing(dose=5000, nbr.doses=14, dosing.interval=12)# maintenance

ev
```

Notice that the units were specified in the table. When specified, the
units use the `units` package to keep track of the units and convert
them if needed.  Additionally, `ggforce` uses them to label the
`ggplot` axes.  The `set_units` and `drop_units` are useful to set and
drop the RxODE event table units.

In this example, you can see the time axes is labeled:

```{r}
rxSolve(m1, ev) %>% plot(C2)
```


If you are more familiar with the NONMEM/RxODE event records, you can
also specify dosing using `et` with the dose elements directly:

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, until = set_units(3, days), ii=12) # loading doses

ev
```

Which gives:

```{r}
rxSolve(m1, ev) %>% plot(C2)
```

This shows how easy creating event tables can be.

### Adding sampling to an event table

If you notice in the above examples, RxODE generated some default
sampling times since there was not any sampling times.  If you wish
more control over the sampling time, you should add the samples to the
RxODE event table by `add.sampling` or `et`

```{r}
ev <- eventTable(amount.units="mg", time.units="hr")

## The methods ar attached to the event table, so you can use them
## directly
ev$add.dosing(dose=10000, nbr.doses = 3)# loading doses

ev$add.sampling(seq(0,24,by=4))

ev
```
Which gives:

```{r}
solve(m1, ev) %>% plot(C2)
```

Or if you use `et` you can simply add them in a similar way to `add.sampling`:

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, until = set_units(3, days), ii=12) %>% # loading doses
    et(seq(0,24,by=4))

ev
```

which gives the following RxODE solve:

```{r}
solve(m1, ev) %>% plot(C2)
```

Note the jagged nature of these plots since there was only a few
sample times.

### Expand the event table to a multi-subject event table.

The only thing that is needed to expand an event table is a list of
IDs that you want to expand;

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, until = set_units(3, days), ii=12) %>% # loading doses
    et(seq(0,48,length.out=200)) %>%
    et(id=1:4)

ev
```

You can see in the following simulation there are 4 individuals that are solved for:

```{r}
set.seed(42)
solve(m1, ev,
      params=data.frame(KA=0.294*exp(rnorm(4)), 18.6*exp(rnorm(4)))) %>%
    plot(C2)
```

### Add doses and samples within a sampling window

In addition to adding fixed doses and fixed sampling times, you can
have windows where you sample and draw doses from.  For dosing windows
you specify the time as an ordered numerical vector with the lowest
dosing time and the highest dosing time inside a list.

In this example, you start with a dosing time with a 6 hour dosing window:

```{r}
set.seed(42)
ev <- et(timeUnits="hr") %>%
    et(time=list(c(0,6)), amt=10000, until = set_units(2, days), ii=12) %>% # loading doses
    et(id=1:4)

ev
```

You can clearly see different dosing times in the following simulation:
```{r}
ev <- ev %>% et(seq(0,48,length.out=200))

solve(m1, ev, params=data.frame(KA=0.294*exp(rnorm(4)), 18.6*exp(rnorm(4)))) %>% plot(C2)
```

Of course in reality the dosing interval may only be 2 hours:

```{r}
set.seed(42)
ev <- et(timeUnits="hr") %>%
    et(time=list(c(0,2)), amt=10000, until = set_units(2, days), ii=12) %>% # loading doses
    et(id=1:4) %>%
    et(seq(0,48,length.out=200))

solve(m1, ev, params=data.frame(KA=0.294*exp(rnorm(4)), 18.6*exp(rnorm(4)))) %>% plot(C2)
```

The same sort of thing can be specified with sampling times.  To
specify the sampling times in terms of a sampling window, you can
create a list of the sampling times.  Each sampling time will be a two
element ordered numeric vector.

```{r}
set.seed(42)
ev <- et(timeUnits="hr") %>%
    et(time=list(c(0,2)), amt=10000, until = set_units(2, days), ii=12) %>% # loading doses
    et(id=1:4)

## Create 20 samples in the first 24 hours and 20 samples in the second 24 hours
samples <- c(lapply(1:20, function(...){c(0,24)}),
             lapply(1:20, function(...){c(20,48)}))

## Add the random collection to the event table
ev <- ev %>% et(samples)

library(ggplot2)
solve(m1, ev, params=data.frame(KA=0.294*exp(rnorm(4)), 18.6*exp(rnorm(4)))) %>% plot(C2) + geom_point()
```

This shows the flexibility in dosing and sampling that the RxODE event tables allow.

### Combining event tables

Since you can create dosing records and sampling records, you can
create any complex dosing regimen you wish.  In addition, RxODE allows
you to combine event tables by `c`, `seq`, `rep`, and `rbind`.

#### Sequencing event tables

One way to combine event table is to sequence them by `c`, `seq` or
`etSeq`. This takes the two dosing groups and adds at least one
inter-dose interval between them:

```{r}
## bid for 5 days
bid <- et(timeUnits="hr") %>%
       et(amt=10000,ii=12,until=set_units(5, "days"))

## qd for 5 days
qd <- et(timeUnits="hr") %>%
      et(amt=20000,ii=24,until=set_units(5, "days"))

## bid for 5 days followed by qd for 5 days
et <- seq(bid,qd) %>% et(seq(0,11*24,length.out=100));

rxSolve(m1, et) %>% plot(C2)
```

When sequencing events, you can also separate this sequence by a
period of time; For example if you wanted to separate this by a week,
you could easily do that with the following sequence of event tables:

```{r}
## bid for 5 days followed by qd for 5 days
et <- seq(bid,set_units(1, "week"), qd) %>%
    et(seq(0,18*24,length.out=100));

rxSolve(m1, et) %>% plot(C2)
```

Note that in this example the time between the bid and the qd event
tables is exactly one week, not 1 week plus 24 hours because of the
inter-dose interval.  If you want that behavior, you can sequence it
using the `wait="+ii"`.

```{r}
## bid for 5 days followed by qd for 5 days
et <- seq(bid,set_units(1, "week"), qd,wait="+ii") %>%
    et(seq(0,18*24,length.out=100));

rxSolve(m1, et) %>% plot(C2)
```

Also note, that RxODE assumes that the dosing is what you want to
space the event tables by, and clears out any sampling records when
you combine the event tables.  If that is not true, you can also use
the option `samples="use"`

#### Repeating event tables
You can have an event table that you can repeat with `etRep` or `rep`.
For example 4 rounds of 2 weeks on QD therapy and 1 week off of
therapy can be simply specified:

```{r}
qd <-et(timeUnits = "hr") %>% et(amt=10000, ii=24, until=set_units(2, "weeks"), cmt="depot")

et <- rep(qd, times=4, wait=set_units(1,"weeks")) %>%
      add.sampling(set_units(seq(0, 12.5,by=0.005),weeks))

rxSolve(m1, et)  %>% plot(C2)
```

This is a simplified way to use a sequence of event tables.
Therefore, many of the same options still apply; That is `samples` are
cleared unless you use `samples="use"`, and the time between event
tables is at least the inter-dose interval.  You can adjust the timing
by the `wait` option.

#### Combining event tables with rbind

You may combine event tables with `rbind`.  This does not consider the
event times when combining the event tables, but keeps them the same
times. If you space the event tables by a waiting period, it also does
not consider the inter-dose interval.

Using the previous `seq` you can clearly see the difference.  Here was the sequence:
```{r}
## bid for 5 days
bid <- et(timeUnits="hr") %>%
       et(amt=10000,ii=12,until=set_units(5, "days"))

## qd for 5 days
qd <- et(timeUnits="hr") %>%
      et(amt=20000,ii=24,until=set_units(5, "days"))

et <- seq(bid,qd) %>%
    et(seq(0,18*24,length.out=500));

rxSolve(m1, et) %>% plot(C2)
```

But if you bind them together with `rbind`
```{r}
## bid for 5 days
et <- rbind(bid,qd) %>%
    et(seq(0,18*24,length.out=500));

rxSolve(m1, et) %>% plot(C2)
```

Still the waiting period applies (but does not consider the inter-dose interval)

```{r}
et <- rbind(bid,wait=set_units(10,days),qd) %>%
    et(seq(0,18*24,length.out=500));

rxSolve(m1, et) %>% plot(C2)
```

You can also bind the tables together and make each ID in the event
table unique; This can be good to combine cohorts with different
expected dosing and sampling times.  This requires the `id="unique"`
option;  Using the first example shows how this is different in this case:

```{r}
## bid for 5 days
et <- etRbind(bid,qd, id="unique") %>%
    et(seq(0,150,length.out=500));

library(ggplot2)
rxSolve(m1, et) %>% plot(C2) + facet_wrap( ~ id)
```

## Event Types for RxODE

### Bolus Doses

A bolus dose is the default type of dose in RxODE and only requires
the `amt`/`dose`

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev) %>% plot(C2) +
    xlab("Time")
```

### Constant Infusion (in terms of duration and rate)

The next type of event is an infusion; There are two ways to specify
an infusion; The first is the `dur` keyword.

An example of this is:

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24, dur=8) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev) %>% plot(depot, C2) +
    xlab("Time")
```

It can be also specified by the `rate` component:

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24, rate=10000/8) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev) %>% plot(depot, C2) +
    xlab("Time")
```

These are the same with the exception of how bioavailability changes
the infusion.

In the case of modeling `rate`, a bioavailability decrease,
decreases the infusion duration, as in NONMEM. For example:

```{r}
rxSolve(m1, ev, c(fdepot=0.25)) %>% plot(depot, C2) +
    xlab("Time")
```

Similarly increasing the bioavailability increases the infusion duration. 

```{r}
rxSolve(m1, ev, c(fdepot=1.25)) %>% plot(depot, C2) +
    xlab("Time")
```

The rationale for this behavior is that the `rate` and `amt` are
specified by the event table, so the only thing that can change with a
bioavailability increase is the duration of the infusion.

If you specify the `amt` and `dur` components in the event table,
bioavailability changes affect the `rate` of infusion.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24, dur=8) %>%
    et(seq(0, 24, length.out=100))

```

You can see the side-by-side comparison of bioavailability changes
affecting `rate` instead of duration with these records in the
following plots:

```{r}
library(ggplot2)
library(gridExtra)

p1 <- rxSolve(m1, ev, c(fdepot=1.25)) %>% plot(depot) +
    xlab("Time") + ylim(0,5000)

p2 <- rxSolve(m1, ev, c(fdepot=0.25)) %>% plot(depot) +
    xlab("Time")+ ylim(0,5000)

grid.arrange(p1,p2, nrow=1)
```

### Modeled Rate and Duration of Infusion

You can model the duration, which is equivalent to NONMEM's
`rate=-2`. As a mnemonic you can use the `dur=model` instead of
`rate=-2`

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24, dur=model) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev, c(durDepot=7)) %>% plot(depot, C2) +
    xlab("Time")

```

Similarly, you may also model rate.  This is equaivlent to NONMEM's
`rate=-1` and is how RxODE's event table specifies the data item as
well. You can also use `rate=model` as a mnemonic:

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12,until=24, rate=model) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev, c(rateDepot=10000/3)) %>% plot(depot, C2) +
    xlab("Time")
```

### Steady State

Steady state doses; These doses are solved until a steady state is
reached with a constant inter-dose interval.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, ss=1) %>%
    et(seq(0, 24, length.out=100))

ev
```

```{r}
rxSolve(m1, ev) %>% plot(C2)
```

### Steady state for complex dosing

By using the `ss=2` flag, you can use the super-positioning principle
in linear kinetics to get steady state nonstandard dosing
(i.e. morning 100 mg vs evening 150 mg). This is done by:

 - Saving all the state values
 - Resetting all the states and solving the system to steady state
 - Adding back all the prior state values
 
```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=24, ss=1) %>%
    et(time=12, amt=15000, ii=24, ss=2) %>%
    et(time=24, amt=10000, ii=24, addl=3) %>%
    et(time=36, amt=15000, ii=24, addl=3) %>%
    et(seq(0, 64, length.out=500))

library(ggplot2)

rxSolve(m1, ev,maxsteps=10000) %>% plot(C2) +
    annotate("rect", xmin=0, xmax=24, ymin=-Inf, ymax=Inf, alpha=0.2) +
    annotate("text", x=12.5, y=7, label="Initial Steady State Period") +
    annotate("text", x=44,   y=7, label="Steady State AM/PM dosing")

```

You can see that it takes a full dose cycle to reach the true complex
steady state dosing.

### Reset Events

Reset events are implemented by `evid=3` or `evid=reset`, for reset
and `evid=4` for reset and dose.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, addl=3) %>%
    et(time=6, evid=reset) %>%
    et(seq(0, 24, length.out=100))

ev
```

The solving show what happens in this system when the system is reset at 6 hours post-dose.

```{r}
rxSolve(m1, ev) %>% plot(depot,C2, eff)
```

You can see all the compartments are reset to their initial values.  The next dose start the dosing cycle over.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, addl=3) %>%
    et(time=6, amt=10000, evid=4) %>%
    et(seq(0, 24, length.out=100))

ev
```

In this case, the whole system is reset and the dose is given

```{r}
rxSolve(m1, ev) %>% plot(depot,C2, eff)
```

### Turning off compartments

You may also turn off a compartment, which is similar to a reset event.
```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, addl=3) %>%
    et(time=6, cmt="-depot", evid=2) %>%
    et(seq(0, 24, length.out=100))

ev
```

Solving shows what this does in the system:

```{r}
rxSolve(m1, ev) %>% plot(depot,C2, eff)
```

In this case, the depot is turned off, and the compartment
concentrations are set to the initial values but the other compartment
concentrations/levels are not reset. When another dose to the depot is
administered the depot compartment is turned back on.

Note that a dose to a compartment only turns back on the compartment
that was dosed.  Hence if you turn off the effect compartment, it
continues to be off after another dose to the depot.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, addl=3) %>%
    et(time=6, cmt="-eff", evid=2) %>%
    et(seq(0, 24, length.out=100))

rxSolve(m1, ev) %>% plot(depot,C2, eff)
```

To turn back on the compartment, a zero-dose to the compartment or a
evid=2 with the compartment would be needed.

```{r}
ev <- et(timeUnits="hr") %>%
    et(amt=10000, ii=12, addl=3) %>%
    et(time=6, cmt="-eff", evid=2) %>%
    et(time=12,cmt="eff",evid=2) %>%
    et(seq(0, 24, length.out=100))

rxSolve(m1, ev) %>% plot(depot,C2, eff)

```

## Classic RxODE evid values

While RxODE still supports these values, this is primarily provided
for historic information, and we recommend using the normal NONMEM
dataset standard that is used by many modeling tools like NONMEM,
Monolix and nlmixr, described above.

Classically, RxODE supported event coding in a single event id `evid` described in the following table.


| 100+ cmt | Infusion Flag                 | <99 Cmt  | SS flag & Turning of Compartment                          |
|----------|-------------------------------|----------|-----------------------------------------------------------|
| 100+ cmt | 0 = bolus dose                | < 99 cmt | 1 = dose                                                  |
|          | 1 = infusion (rate)           |          | 10 = Steady state 1 (equivalent to SS=1)                  |
|          | 2 = infusion (dur)            |          | 20 = Steady state 2 (equivalent to SS=2)                  |
|          | 6 = turn off modeled duration |          | 30 = Turn off a compartment (equivalent to -CMT w/EVID=2) |
|          | 7 = turn off modeled rate     |          |                                                           |
|          | 8 = turn on modeled duration  |          |                                                           |
|          | 9 = turn on modeled rate      |          |                                                           |

The classic EVID concatenate the numbers in the above table, so an
infusion would to compartment 1 would be `10101` and an infusion to compartment 199 would be `119901`.

EVID = 0 (observations), EVID=2 (other type event) and EVID=3 are all
supported.  EVID 10-99 represents modeled time interventions, similar
to NONMEM's MTIME. This along with amount (amt) and time columns specify the events in the ODE system.

For infusions specified with EVIDs > 100 the amt column represents the rate value.

For Infusion flags 1 and 2 `+amt` turn on the infusion to a specific
compartment `-amt` turn off the infusion to a specific
compartment. To specify a dose/duration you place the dosing records
at the time the duration starts or stops.

For modeled rate/duration infusion flags the on infusion flag must be followed by an off infusion record.

These number are concatenated together to form a full RxODE event ID, as shown in the following examples:

### Bolus Dose Examples
*A 100 bolus dose to compartment #1 at time 0*

| time | evid | amt |
|------|------|-----|
| 0    | 101  | 100 |
| 0.5  | 0    | 0   |
| 1    | 0    | 0   |


*A 100 bolus dose to compartment #99 at time 0*

| time | evid | amt |
|------|------|-----|
| 0    | 9901 | 100 |
| 0.5  | 0    | 0   |
| 1    | 0    | 0   |

*A 100 bolus dose to compartment #199 at time 0*

| time | evid   | amt |
|------|--------|-----|
| 0    | 109901 | 100 |
| 0.5  | 0      | 0   |
| 1    | 0      | 0   |

### Infusion Event Examples

Bolus infusion with rate 50 to compartment 1 for 1.5 hr, (modeled
	bioavailability changes duration of infusion)

| time | evid  | amt |
|------|-------|-----|
| 0    | 10101 | 50  |
| 0.5  | 0     | 0   |
| 1    | 0     | 0   |
| 1.5  | 10101 | -50 |


Bolus infusion with rate 50 to compartment 1 for 1.5 hr (modeled
	bioavailability changes rate of infusion)

| time | evid  | amt |
|------|-------|-----|
| 0    | 20101 | 50  |
| 0.5  | 0     | 0   |
| 1    | 0     | 0   |
| 1.5  | 20101 | -50 |

Modeled rate with amount of 50


| time | evid  | amt |
|------|-------|-----|
| 0    | 90101 | 50  |
| 0    | 70101 | 50  |
| 0.5  | 0     | 0   |
| 1    | 0     | 0   |


Modeled duration with amount of 50

| time | evid  | amt |
|------|-------|-----|
| 0    | 80101 | 50  |
| 0    | 60101 | 50  |
| 0.5  | 0     | 0   |
| 1    | 0     | 0   |

### Steady State for classic RxODE EVID example

Steady state dose to cmt 1

| time | evid | amt |
|------|------|-----|
| 0    | 110  | 50  |

Steady State with super-positioning principle for am 50 and pm 100 dose

| time | evid | amt |
|------|------|-----|
| 0    | 110  | 50  |
| 12   | 120  | 100 |

### Turning off a compartment with classic RxODE EVID

Turn off the first compartment at time 12

| time | evid | amt |
|------|------|-----|
| 0    | 110  | 50  |
| 12   | 130  | NA  |




Event coding in RxODE is encoded in a single event number `evid`. For
compartments under 100, this is coded as:


- This event is `0` for observation events.
- For a specified compartment a bolus dose is defined as:
  - 100\*(Compartment Number) + 1
  - The dose is then captured in the `amt`
- For IV bolus doses the event is defined as:
   - 10000 + 100\*(Compartment Number) + 1
   - The infusion rate is captured in the `amt` column
   - The infusion is turned off by subtracting `amt` with the same
     `evid` at the stop of the infusion.
	 
	 
For compartments greater or equal to 100, the 100s place and above
digits are transfered to the 100,000th place digit. For doses to the
99th compartment the `evid` for a bolus dose would be `9901` and the
`evid` for an infusion would be `19901`.  For a bolus dose to the
`199`th compartment the `evid` for the bolus dose would be
`109901`. An infusion dosing record for the `199`th compartment would
be `119901`.

