---
title: "RxODE Transit Compartment Models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RxODE Transit Compartment Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
```

Savic 2008 first introduced the idea of transit compartments being a
mechanistic explanation of a a lag-time type phenomena. RxODE has special handling of these models:

You can specify this in a similar manner as the original paper:

```{r out.width="100%"}
library(RxODE)
library(ggplot2)
mod <- RxODE({
    ## Table 3 from Savic 2007
    cl = 17.2 # (L/hr)
    vc = 45.1 # L
    ka = 0.38 # 1/hr
    mtt = 0.37 # hr
    bio=1
    n = 20.1
    k = cl/vc
    ktr = (n+1)/mtt
    ## note that lgammafn is the same as lgamma in R.
    d/dt(depot) = exp(log(bio*podo)+log(ktr)+n*log(ktr*t)-ktr*t-lgammafn(n+1))-ka*depot
    d/dt(cen) = ka*depot-k*cen
})

et <- eventTable();
et$add.sampling(seq(0, 7, length.out=200));
et$add.dosing(20, start.time=0);

transit <- rxSolve(mod, et, transit_abs=TRUE)

transit %>% ggplot(aes(time,cen))+geom_line()+ylab("Central Concentration")+
xlab("")
```


Another option is to specify the transit compartment function `transit` syntax.  This specifies the parameters
`transit(number of transit compartments, mean transit time, bioavailibility)`.  The bioavailibity term is optional.

Using the `transit` code also automatically turns on the `transit_abs`
option.  Therefore, the same model can be specified by:


```{r}

mod <- RxODE({
    ## Table 3 from Savic 2007
    cl = 17.2 # (L/hr)
    vc = 45.1 # L
    ka = 0.38 # 1/hr
    mtt = 0.37 # hr
    bio=1
    n = 20.1
    k = cl/vc
    ktr = (n+1)/mtt
    d/dt(depot) = transit(n,mtt,bio)-ka*depot
    d/dt(cen) = ka*depot-k*cen
})

et <- eventTable();
et$add.sampling(seq(0, 7, length.out=200));
et$add.dosing(20, start.time=0);

transit <- rxSolve(mod, et)

transit %>% ggplot(aes(time,cen))+geom_line()+ylab("Central Concentration") +
xlab("")
```

