---
title: "Plotting RxODE models"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
---

## Plotting in RxODE

### Setting up model for plotting explanation

The first step to explain the RxODE model plots is to setup:

1. An RxODE model
2. An event table
3. A solved object

```{r}
library(RxODE)
## Model from RxODE tutorial
m1 <- RxODE({
  KA <- 2.94E-01
  CL <- 1.86E+01
  V2 <- 4.02E+01
  Q <- 1.05E+01
  V3 <- 2.97E+02
  Kin <- 1
  Kout <- 1
  EC50 <- 200
  ## Added modeled bioavaiblity, duration and rate
  fdepot <- 1
  durDepot <- 8
  rateDepot <- 1250
  C2 <- centr / V2
  C3 <- peri / V3
  d/dt(depot) <- -KA * depot
  f(depot) <- fdepot
  dur(depot) <- durDepot
  rate(depot) <- rateDepot
  d/dt(centr) <- KA * depot - CL * C2 - Q * C2 + Q * C3
  d/dt(peri) <- Q * C2 - Q * C3
  d/dt(eff) <- Kin - Kout * (1 - C2 / (EC50 + C2)) * eff
  eff(0) <- 1
})

ev <- et(timeUnits = "hr") %>%
  et(amt = 10000, ii = 12, until = 24) %>%
  et(seq(0, 24, length.out = 100))

s <- rxSolve(m1, ev)
```

### Default RxODE plot

The default plot in `RxODE` is to plot time versus `state` or calculated `lhs` values:

```{r}
plot(s)
```

Things to note:

First, these plots are actually `ggplot2` plots and can be modified by
the standard `ggplot2` grammar of graphics

Also note that the `plot` method for `rxSolve` objects can currently
use some of the arguments of the default plot method:

- `log` a character string which contains `"x"` if the x axis is to be
  logarithmic, `"y"` if the y axis is to be logarithmic and `"xy"` or
  `"yx"` if both axes are to be logarithmic. If available, this will
  use the `xgxr` `xgx_scale_x_log10()` and `xgx_scale_y_log10()`
  functions.

- `xlab` This is a bit different than simply supplying the x-label
  through `ggplot2`, it will retain the unit information originally
  supplied in the event table.  Also note, if available, this unit
  information is used will be used to have more meaningful ticks with
  `xgx_scale_x_time_units()`

- `ylab` This supplies the y label

The following are currently unsupported:
   a. `type`
   b. `xlim`
   c. `ylim`
   d. `main`
   e. `sub`
   f. `ann`
   g. `axes`


## Subsetting plots 

While this is very interesting, it is often more useful to subset
plots based on values you are more interested in; For example, it is
more interesting in this plot to `eff` and `C2`.  To do this, you
simply specify the interesting items after the plot function.  For instance:

```{r}
plot(s, C2, eff)
```

## Semi-log plots

Semi-log plots of PK concentrations are very common; To do this you
simply need to use `log="y"`

```{r}
plot(s, C2, log="y", ylab="Concentration")
```

This uses the `xgx_scale_y_log10()` when available to more clearly
show the semi-log nature of the plot.

## Plots with multi-subject plots

If you have multi-subject plots you can easily plot applying the same
principles as above. To illustrate this plot, lets expand an event table to 
include a plot of 4 subjects with lognormal random variability between subjects;

```{r}
# Setup the new problem
m2 <- RxODE({
  KA <- 2.94E-01
  CL <- 1.86E+01 * exp(eta.Cl)
  V2 <- 4.02E+01
  Q <- 1.05E+01
  V3 <- 2.97E+02
  Kin <- 1
  Kout <- 1
  EC50 <- 200
  ## Added modeled bioavaiblity, duration and rate
  fdepot <- 1
  durDepot <- 8
  rateDepot <- 1250
  C2 <- centr / V2
  C3 <- peri / V3
  d / dt(depot) <- -KA * depot
  f(depot) <- fdepot
  dur(depot) <- durDepot
  rate(depot) <- rateDepot
  d / dt(centr) <- KA * depot - CL * C2 - Q * C2 + Q * C3
  d / dt(peri) <- Q * C2 - Q * C3
  d / dt(eff) <- Kin - Kout * (1 - C2 / (EC50 + C2)) * eff
  eff(0) <- 1
})

# Add the between subject variability:
omega <- lotri(eta.Cl ~ 0.4^2)

# Create the event table
ev <- et(timeUnits = "hr") %>%
  et(amt = 10000, until = units::set_units(3, days), ii = 12) %>% # loading doses
  et(seq(0, 48, length.out = 200)) %>%
  et(id = 1:4)

s <- rxSolve(m2, ev, omega=omega)
```

Once that is complete, you may plot it by the same method:

```{r}
plot(s, C2, eff)
```

Notice that this is colored by each individual and labeled with a legend.

If you are only interested in the concentration, it produces a similar plot:

```{r}
plot(s, C2, log="y", ylab="Concentration")
```

Notice that this plot each individual is labeled by an attached id to
let you know where the individual comes from.  This is done by
`ggrepel` if it is available, otherwise a legend is retained.

### Multi-subject plots with large number of subjects 

This can get a bit much when there are many subjects that are solved
(in this example lets simulate 100)


```{r}
ev <- et(timeUnits = "hr") %>%
  et(amt = 10000, until = units::set_units(3, days), ii = 12) %>% # loading doses
  et(seq(0, 48, length.out = 200)) %>%
  et(id = 1:100) # 100 subjects

s <- rxSolve(m2, ev, omega=omega)

plot(s, C2, log="y", ylab="Concentration")
```

In this case, all the individuals are put on the plot in transparent
greyscale and plot on the same pane.  This allows the places where
more subjects are present to be darker.


The number of individuals where the plots switch from legend to
grayscale is controlled by changing the options for `RxODE.spaghetti`.
ie. `options(RxODE.spaghetti=7)`.  7 individuals is the default value
when the plotting changes from individual to grayscale spaghetti
plots.

You can also create a confidence interval of these simulations with `confint`:

```{r}
s2 <- confint(s, parm="C2")
```

And plot this with `plot`

```{r}
plot(s2)
```
