---
title: "Changing RxODE compartment numbers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing RxODE compartment numbers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(cli.unicode=FALSE, crayon.enabled=FALSE);
options(knitr.table.format = "html")
knitr::opts_chunk$set( comment = "#>")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
options(width=80)
Sys.setenv(RSTUDIO_CONSOLE_WIDTH=80)
```

# How RxODE assigns compartment numbers

RxODE automatically assigns compartment numbers when parsing. For
example, with the Mavoglurant PBPK model the following model may be
used:

```{r}
library(RxODE)
pbpk <- RxODE({
    KbBR = exp(lKbBR)
    KbMU = exp(lKbMU)
    KbAD = exp(lKbAD)
    CLint= exp(lCLint + eta.LClint)
    KbBO = exp(lKbBO)
    KbRB = exp(lKbRB)

    ## Regional blood flows
    CO  = (187.00*WT^0.81)*60/1000;         # Cardiac output (L/h) from White et al (1968)
    QHT = 4.0 *CO/100;
    QBR = 12.0*CO/100;
    QMU = 17.0*CO/100;
    QAD = 5.0 *CO/100;
    QSK = 5.0 *CO/100;
    QSP = 3.0 *CO/100;
    QPA = 1.0 *CO/100;
    QLI = 25.5*CO/100;
    QST = 1.0 *CO/100;
    QGU = 14.0*CO/100;
    QHA = QLI - (QSP + QPA + QST + QGU); # Hepatic artery blood flow
    QBO = 5.0 *CO/100;
    QKI = 19.0*CO/100;
    QRB = CO - (QHT + QBR + QMU + QAD + QSK + QLI + QBO + QKI);
    QLU = QHT + QBR + QMU + QAD + QSK + QLI + QBO + QKI + QRB;

    ## Organs' volumes = organs' weights / organs' density
    VLU = (0.76 *WT/100)/1.051;
    VHT = (0.47 *WT/100)/1.030;
    VBR = (2.00 *WT/100)/1.036;
    VMU = (40.00*WT/100)/1.041;
    VAD = (21.42*WT/100)/0.916;
    VSK = (3.71 *WT/100)/1.116;
    VSP = (0.26 *WT/100)/1.054;
    VPA = (0.14 *WT/100)/1.045;
    VLI = (2.57 *WT/100)/1.040;
    VST = (0.21 *WT/100)/1.050;
    VGU = (1.44 *WT/100)/1.043;
    VBO = (14.29*WT/100)/1.990;
    VKI = (0.44 *WT/100)/1.050;
    VAB = (2.81 *WT/100)/1.040;
    VVB = (5.62 *WT/100)/1.040;
    VRB = (3.86 *WT/100)/1.040;

    ## Fixed parameters
    BP = 0.61;      # Blood:plasma partition coefficient
    fup = 0.028;    # Fraction unbound in plasma
    fub = fup/BP;   # Fraction unbound in blood

    KbLU = exp(0.8334);
    KbHT = exp(1.1205);
    KbSK = exp(-.5238);
    KbSP = exp(0.3224);
    KbPA = exp(0.3224);
    KbLI = exp(1.7604);
    KbST = exp(0.3224);
    KbGU = exp(1.2026);
    KbKI = exp(1.3171);


    ##-----------------------------------------
    S15 = VVB*BP/1000;
    C15 = Venous_Blood/S15

    ##-----------------------------------------
    d/dt(Lungs) = QLU*(Venous_Blood/VVB - Lungs/KbLU/VLU);
    d/dt(Heart) = QHT*(Arterial_Blood/VAB - Heart/KbHT/VHT);
    d/dt(Brain) = QBR*(Arterial_Blood/VAB - Brain/KbBR/VBR);
    d/dt(Muscles) = QMU*(Arterial_Blood/VAB - Muscles/KbMU/VMU);
    d/dt(Adipose) = QAD*(Arterial_Blood/VAB - Adipose/KbAD/VAD);
    d/dt(Skin) = QSK*(Arterial_Blood/VAB - Skin/KbSK/VSK);
    d/dt(Spleen) = QSP*(Arterial_Blood/VAB - Spleen/KbSP/VSP);
    d/dt(Pancreas) = QPA*(Arterial_Blood/VAB - Pancreas/KbPA/VPA);
    d/dt(Liver) = QHA*Arterial_Blood/VAB + QSP*Spleen/KbSP/VSP + QPA*Pancreas/KbPA/VPA + QST*Stomach/KbST/VST + QGU*Gut/KbGU/VGU - CLint*fub*Liver/KbLI/VLI - QLI*Liver/KbLI/VLI;
    d/dt(Stomach) = QST*(Arterial_Blood/VAB - Stomach/KbST/VST);
    d/dt(Gut) = QGU*(Arterial_Blood/VAB - Gut/KbGU/VGU);
    d/dt(Bones) = QBO*(Arterial_Blood/VAB - Bones/KbBO/VBO);
    d/dt(Kidneys) = QKI*(Arterial_Blood/VAB - Kidneys/KbKI/VKI);
    d/dt(Arterial_Blood) = QLU*(Lungs/KbLU/VLU - Arterial_Blood/VAB);
    d/dt(Venous_Blood) = QHT*Heart/KbHT/VHT + QBR*Brain/KbBR/VBR + QMU*Muscles/KbMU/VMU + QAD*Adipose/KbAD/VAD + QSK*Skin/KbSK/VSK + QLI*Liver/KbLI/VLI + QBO*Bones/KbBO/VBO + QKI*Kidneys/KbKI/VKI + QRB*Rest_of_Body/KbRB/VRB - QLU*Venous_Blood/VVB;
    d/dt(Rest_of_Body) = QRB*(Arterial_Blood/VAB - Rest_of_Body/KbRB/VRB);
})

```

If you look at the summary, you can see where RxODE assigned the compartment number(s)

```{r}
summary(pbpk)
```

In this case, it is assigned to compartment `15`.  Figuring this out
can be inconvenient and also lead to re-numbering compartment in
simulation or estimation datasets. While it is easy and probably
clearer to specify the [compartment by name](RxODE-events.html), other
tools only support compartment numbers.  Therefore, having a way to number
compartment easily can lead to less data modification between multiple
tools.

# Changing compartment numbers by pre-declaring the compartments

To add the compartments to the RxODE model in the order you desire you
simply need to pre-declare the compartments with `cmt`.  For example
specifying is `Venous_Blood` and `Skin` to be the 1st and 2nd
compartments, respectively, is simple:

```{r}
pbpk2 <- RxODE({
    cmt(Venous_Blood) ## Now this is the first compartment, ie cmt=1
    cmt(Skin) ## Skin may be a compartment you wish to dose to as well, so it is now cmt=2
    KbBR = exp(lKbBR)
    KbMU = exp(lKbMU)
    KbAD = exp(lKbAD)
    CLint= exp(lCLint + eta.LClint)
    KbBO = exp(lKbBO)
    KbRB = exp(lKbRB)

    ## Regional blood flows
    CO  = (187.00*WT^0.81)*60/1000;         # Cardiac output (L/h) from White et al (1968)
    QHT = 4.0 *CO/100;
    QBR = 12.0*CO/100;
    QMU = 17.0*CO/100;
    QAD = 5.0 *CO/100;
    QSK = 5.0 *CO/100;
    QSP = 3.0 *CO/100;
    QPA = 1.0 *CO/100;
    QLI = 25.5*CO/100;
    QST = 1.0 *CO/100;
    QGU = 14.0*CO/100;
    QHA = QLI - (QSP + QPA + QST + QGU); # Hepatic artery blood flow
    QBO = 5.0 *CO/100;
    QKI = 19.0*CO/100;
    QRB = CO - (QHT + QBR + QMU + QAD + QSK + QLI + QBO + QKI);
    QLU = QHT + QBR + QMU + QAD + QSK + QLI + QBO + QKI + QRB;

    ## Organs' volumes = organs' weights / organs' density
    VLU = (0.76 *WT/100)/1.051;
    VHT = (0.47 *WT/100)/1.030;
    VBR = (2.00 *WT/100)/1.036;
    VMU = (40.00*WT/100)/1.041;
    VAD = (21.42*WT/100)/0.916;
    VSK = (3.71 *WT/100)/1.116;
    VSP = (0.26 *WT/100)/1.054;
    VPA = (0.14 *WT/100)/1.045;
    VLI = (2.57 *WT/100)/1.040;
    VST = (0.21 *WT/100)/1.050;
    VGU = (1.44 *WT/100)/1.043;
    VBO = (14.29*WT/100)/1.990;
    VKI = (0.44 *WT/100)/1.050;
    VAB = (2.81 *WT/100)/1.040;
    VVB = (5.62 *WT/100)/1.040;
    VRB = (3.86 *WT/100)/1.040;

    ## Fixed parameters
    BP = 0.61;      # Blood:plasma partition coefficient
    fup = 0.028;    # Fraction unbound in plasma
    fub = fup/BP;   # Fraction unbound in blood

    KbLU = exp(0.8334);
    KbHT = exp(1.1205);
    KbSK = exp(-.5238);
    KbSP = exp(0.3224);
    KbPA = exp(0.3224);
    KbLI = exp(1.7604);
    KbST = exp(0.3224);
    KbGU = exp(1.2026);
    KbKI = exp(1.3171);


    ##-----------------------------------------
    S15 = VVB*BP/1000;
    C15 = Venous_Blood/S15

    ##-----------------------------------------
    d/dt(Lungs) = QLU*(Venous_Blood/VVB - Lungs/KbLU/VLU);
    d/dt(Heart) = QHT*(Arterial_Blood/VAB - Heart/KbHT/VHT);
    d/dt(Brain) = QBR*(Arterial_Blood/VAB - Brain/KbBR/VBR);
    d/dt(Muscles) = QMU*(Arterial_Blood/VAB - Muscles/KbMU/VMU);
    d/dt(Adipose) = QAD*(Arterial_Blood/VAB - Adipose/KbAD/VAD);
    d/dt(Skin) = QSK*(Arterial_Blood/VAB - Skin/KbSK/VSK);
    d/dt(Spleen) = QSP*(Arterial_Blood/VAB - Spleen/KbSP/VSP);
    d/dt(Pancreas) = QPA*(Arterial_Blood/VAB - Pancreas/KbPA/VPA);
    d/dt(Liver) = QHA*Arterial_Blood/VAB + QSP*Spleen/KbSP/VSP + QPA*Pancreas/KbPA/VPA + QST*Stomach/KbST/VST + QGU*Gut/KbGU/VGU - CLint*fub*Liver/KbLI/VLI - QLI*Liver/KbLI/VLI;
    d/dt(Stomach) = QST*(Arterial_Blood/VAB - Stomach/KbST/VST);
    d/dt(Gut) = QGU*(Arterial_Blood/VAB - Gut/KbGU/VGU);
    d/dt(Bones) = QBO*(Arterial_Blood/VAB - Bones/KbBO/VBO);
    d/dt(Kidneys) = QKI*(Arterial_Blood/VAB - Kidneys/KbKI/VKI);
    d/dt(Arterial_Blood) = QLU*(Lungs/KbLU/VLU - Arterial_Blood/VAB);
    d/dt(Venous_Blood) = QHT*Heart/KbHT/VHT + QBR*Brain/KbBR/VBR + QMU*Muscles/KbMU/VMU + QAD*Adipose/KbAD/VAD + QSK*Skin/KbSK/VSK + QLI*Liver/KbLI/VLI + QBO*Bones/KbBO/VBO + QKI*Kidneys/KbKI/VKI + QRB*Rest_of_Body/KbRB/VRB - QLU*Venous_Blood/VVB;
    d/dt(Rest_of_Body) = QRB*(Arterial_Blood/VAB - Rest_of_Body/KbRB/VRB);
})
```

You can see this change in the simple printout

```{r}
pbpk2
```

The first two compartments are `Venous_Blood` followed by `Skin`.

# Appending compartments to the model

You can also append "compartments" to the model.  Because of the ODE
solving internals, you cannot add fake compartments to the model until
after all the differential equations are defined.

For example this is legal:

```{r}
ode.1c.ka <- RxODE({
    C2 = center/V;
    d / dt(depot) = -KA * depot
    d/dt(center) = KA * depot - CL*C2
    cmt(eff);
})
print(ode.1c.ka)
```

Where as this is not:

```
ode.1c.ka <- RxODE({
    cmt(eff);
    C2 = center/V;
    d / dt(depot) = -KA * depot
    d/dt(center) = KA * depot - CL*C2
})
```

Which gives an error:

```
Error in rxModelVars_(obj) : 
  Evaluation error: Compartment 'eff' needs differential equations defined.
```
