---
title: "Introduction to RxODE"
author: "Matthew L. Fidler, Melissa Hallow, and Wenping Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RxODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## RxODE: A tool for performing simulations from Ordinary Differential Equation (ODE) models, with applications for pharmacometrics
***  

##### Authors: Matthew L. Fidler, Melissa Hallow, and Wenping Wang

***

#### Introduction
`RxODE` is an R package that facilitates simulation with ODE models in
R. It is designed with pharmacometrics models in mind, but can be
applied more generally to any ODE model.

***
#### Description of RxODE illustrated through an example
The model equations are specified through a text string in R. Both
differential and algebraic equations are permitted. Differential
equations are specified by `d/dt(var_name) = `. Each
equation can be separated by a semicolon.

```{r}
ode <- "
   C2 = centr/V2;
   C3 = peri/V3;
   d/dt(depot) =-KA*depot;
   d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
   d/dt(peri)  =                    Q*C2 - Q*C3;
   d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
"
```
To load `RxODE` package and compile the model: 
```{r}
library(RxODE)
work <- tempfile("Rx_intro-")
mod1 <- RxODE(model = ode, modName = "mod1", wd = work)
```
##### Specify ODE parameters and initial conditions

Model parameters can be defined as named vectors. Names of parameters in
the vector must be a superset of parameters in the ODE model, and the
order of parameters within the vector is not important. 
```{r}
theta <- 
   c(KA=2.94E-01, CL=1.86E+01, V2=4.02E+01, # central 
     Q=1.05E+01,  V3=2.97E+02,              # peripheral
     Kin=1, Kout=1, EC50=200)               # effects
```

Initial conditions (ICs) are defined through a vector as well. If the
vector is not named, the number of ICs must equal exactly the number of
ODEs in the model, and the order must be the same as the order in
which the ODEs are listed in the model. 

```{r}
inits <- c(0, 0, 0, 1)
```

When elements are named, missing elements are added and set to
zero. Also when named, the order of initilizations does not matter.
Therefore the following code is an equvalent initialization as the
code above.


```{r}
inits <- c(eff=1);
```

##### Specify Dosing and sampling in RxODE

`RxODE` provides a simple and very flexible way to specify dosing and
sampling through functions that generate an event table. First, an
empty event table is generated through the "eventTable()" function:

```{r}
ev <- eventTable(amount.units='mg', time.units='hours')
```

Next, use the `add.dosing()` and `add.sampling()` functions of the
`EventTable` object to specify the dosing (amounts, frequency and/or
times, etc.) and observation times at which to sample the state of the
system.  These functions can be called multiple times to specify more
complex dosing or sampling regiments.  Here, these functions are used
to specify 10mg BID dosing for 5 days, followed by 20mg QD dosing for
5 days:

```{r}
ev$add.dosing(dose=10000, nbr.doses=10, dosing.interval=12)
ev$add.dosing(dose=20000, nbr.doses=5, start.time=120, dosing.interval=24)
ev$add.sampling(0:240)
```

If you wish you can also do this with the `mattigr` pipe operator `%>%`

```{r}
ev <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=10, dosing.interval=12) %>%
    add.dosing(dose=20000, nbr.doses=5, start.time=120,dosing.interval=24) %>%
    add.sampling(0:240);
```

The functions `get.dosing()` and `get.sampling()` can be used to
retrieve information from the event table.

```{r}
head(ev$get.dosing())
```

```{r}
head(ev$get.sampling())
```

##### Solving ODEs

The ODE can now be solved by calling the model object's `run` or `solve`
function. Simulation results for all variables in the model are stored
in the output matrix x. 

```{r}
x <- mod1$solve(theta, ev, inits)
head(x)
```

This returns a matrix.  You can see the compartment values in the plot
below:

```{r}
par(mfrow=c(1,2))
matplot(x[,"C2"], type="l", ylab="Central Concentration")
matplot(x[,"eff"], type="l", ylab = "Effect")
```

##### Using RxODE data frames

You can also return a solved object that is a modified data-frame.
This is done by the `predict()` or `solve()` methods:

```{r}
x <- predict(mod1,theta, ev, inits)
print(x)
```
or

```{r}
x <- solve(mod1,theta, ev, inits)
print(x)
```

Or with `mattigr`

```{r}
x <- mod1 %>% solve(theta, ev, inits)
print(x)
```

The solved object acts as a `data.frame` or `tbl` that can be filtered
by `dpylr`.  For example you could filter it easily.

```{r}
library(dplyr)
x <- mod1 %>% solve(theta,ev,inits) %>%  filter(time <=3)
x
```

However it isn't just a simple data object.  You can use the solved
object to update paramters on the fly, or even change the sampling
time.


First we need to recreate the original solved system:

```{r}
x <- mod1 %>% solve(theta,ev,inits);
```

To examine or change initial conditions, you can use the syntax
`cmt.0`, `cmt0`, or `cmt_0`. In the case of the `eff` compartment
defined by the model, this is:

```{r}
x$eff0
```

which shows the initial condition of the effect compartment.  If you
wished to change this initial condition to 2, this can be done easily
by:

```{r}
x$eff0 <- 2
x
```

Notice that the inital effect is now `2`.

You can also change the sampling times easily by this method by
changing `t` or `time`.  For example:

```{r}
x$t <- seq(0,5,length.out=20)
x
```

You can also access or change parameters by the `$` operator.  For
example, accessing `KA` can be done by:

```{r}
x$KA
```

And you may change it by assigning it to a new value.

```{r}
x$KA <- 1;
x
```


You can access/change all the parametrs, initilizations or events with
the `$params`, `$inits`, `$events` accessor syntax, similar to what is
used above.

This syntax makes it easy to update and explore the effect of various
parameters on the solved object.

#### Mixing Solved Systems and ODEs.

In addition to pure ODEs, you may mix solved systems and ODEs.  The
prior 2-compartment indirect response model can be simplified with a
`linCmt()` function:

```{r}
mod2 <- RxODE({
    ## the order of variables do not matter, the type of compartmental
    ## model is determined by the parameters specified.
    C2   = linCmt(KA, CL, V2, Q, V3);
    eff(0) = 1  ## This specifies that the effect compartment starts at 1.
    d/dt(eff) =  Kin - Kout*(1-C2/(EC50+C2))*eff;
})
```

Like a Sherlock Holmes on the case of a mystery, the `linCmt()`
function figures out the type of model to use based on the parameter
names specified.

Most often, pharmacometric models are parameterized in terms of volume
and clearances. Clearances are specified by NONMEM-style names of
`CL`, `Q`, `Q1`, `Q2`, etc. or distributional clearances `CLD`,
`CLD2`. Volumes are specified by Central (`VC` or `V`),
Peripheral/Tissue (`VP`, `VT`).

Another popular parameterization is in terms of micro-constants. RxODE assumes
compartment `1` is the central compartment.  The elimination constant
would be specified by `K`, `Ke` or `Kel`.

Once the `linCmt()` sleuthing is complete, the `1`, `2` or `3`
compartment model solution is used as the value of `linCmt()`.

This allows the indirect response model above to assign the
2-compartment model to the `C2` variable and the used in the indirect
response model.

When mixing the solved systems and the ODEs, the solved system's
compartment is always the last compartment.  This is because the
solved system technically isn't a compartment to be solved.  Adding
the dosing compartment to the end will not interfere with the actual
ODE to be solved.

Therefore,in the two-compartment indirect response model, the effect
compartment is compartment #1 while the PK dosing compartment for the
depot is compartment #2.

This compartment model requires a new event table since the
compartment number changed:

```{r}
ev <- eventTable(amount.units='mg', time.units='hours') %>%
    add.dosing(dose=10000, nbr.doses=10, dosing.interval=12,dosing.to=2) %>%
    add.dosing(dose=20000, nbr.doses=5, start.time=120,dosing.interval=24,dosing.to=2) %>%
    add.sampling(0:240);
```

This can be solved with the following command:

```{r}
(x <- mod2 %>%  solve(theta, ev))
```

Note this solving did not require specifying the effect compartment
initial condition to be `1`.  Rather, this is already pre-specified by `eff(0)=1`.

This can be solved for different initial conditions easily:

```{r}
(x <- mod2 %>%  solve(theta, ev,c(eff=2)))
```

The RxODE detective also does not require you to specify the variables
in the `linCmt()` function if they are already defined in the block.

Therefore, the following function will also work to solve the same
system.

```{r}
mod3 <- RxODE({
    KA=2.94E-01;
    CL=1.86E+01; 
    V2=4.02E+01; 
    Q=1.05E+01;
    V3=2.97E+02; 
    Kin=1;
    Kout=1;
    EC50=200;
    ## The linCmt() picks up the variables from above
    C2   = linCmt();
    eff(0) = 1  ## This specifies that the effect compartment starts at 1.
    d/dt(eff) =  Kin - Kout*(1-C2/(EC50+C2))*eff;
})

(x <- mod3 %>%  solve(ev))
```

Note that you do not specify the parameters when solving the system
since they are built into the model, but you can override the parameters:

```{r}
(x <- mod3 %>%  solve(c(KA=10),ev))
```

#### ODEs and covariates

Covariates are easy to specify in RxODE, you can specify them as a
variable. Time-varying covariates, like clock time in a circadian
rhythm model, can also be used.  Extending the indirect response model
already discussed, we have:


```{r}
mod3 <- RxODE({
    KA=2.94E-01;
    CL=1.86E+01; 
    V2=4.02E+01; 
    Q=1.05E+01;
    V3=2.97E+02; 
    Kin0=1;
    Kout=1;
    EC50=200;
    ## The linCmt() picks up the variables from above
    C2   = linCmt();
    Tz= 8
    amp=0.1
    eff(0) = 1  ## This specifies that the effect compartment starts at 1.
    ## Kin changes based on time of day (like cortosol)
    Kin =   Kin0 +amp *cos(2*pi*(ctime-Tz)/24)
    d/dt(eff) =  Kin - Kout*(1-C2/(EC50+C2))*eff;
})


ev <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=100));


 ## Create data frame of  8 am dosing for the first dose
cov.df  <- data.frame(ctime =(seq(0,48,length.out=100)+8) %% 24);

```

Now there is a covariate present, the system can be solved using the cov option

```{r}

(r1 <- solve(mod3, ev, covs=cov.df,covs_interpolation="linear"))
```

When solving ODE equations, the solver may sample times outside of the
data.  When this happens, this ODE solver uses linear interpolation
between the covariate values. This is the default value.  It is
equivalent to R's `approxfun` with `method="linear"`, which is the
default `approxfun`.

```{r out.width="100%"}
par(mfrow=c(1,2))
matplot(r1[,"C2"], type="l", ylab="Central Concentration")
matplot(r1[,"eff"], type="l", ylab = "Effect")
```

Note that the linear approximation in this case leads to some kinks in
the solved system at 24-hours where the covariate has a linear
interpolation between near 24 and near 0.

In RxODE, covariate interpolation can also be the last observation
carried forward, or constant approximation.  This is equivalent to R's
`approxfun` with `method="constant"`.


```{r}
(r2 <- solve(mod3, ev, covs=cov.df,covs_interpolation="constant"))
```

which gives the following plots:

```{r out.width="100%"}
par(mfrow=c(1,2))
matplot(r2[,"C2"], type="l", ylab="Central Concentration")
matplot(r2[,"eff"], type="l", ylab = "Effect")
```

In this case, the plots seem to be smoother.

#### RxODE and transit compartment models

Savic 2008 first introduced the idea of transit compartments being a
mechanistic explanation of a a lag-time type phenomena. RxODE has special handling of these models:

You can specify this in a similar manner as the original paper:

```{r out.width="100%"}

mod <- RxODE({
    ## Table 3 from Savic 2007
    cl = 17.2 # (L/hr)
    vc = 45.1 # L
    ka = 0.38 # 1/hr
    mtt = 0.37 # hr
    bio=1
    n = 20.1
    k = cl/vc
    ktr = (n+1)/mtt
    ## note that lgammafn is the same as lgamma in R.
    d/dt(depot) = exp(log(bio*podo)+log(ktr)+n*log(ktr*t)-ktr*t-lgammafn(n+1))-ka*depot
    d/dt(cen) = ka*depot-k*cen
})

et <- eventTable();
et$add.sampling(seq(0, 7, length.out=200));
et$add.dosing(20, start.time=0);

transit <- rxSolve(mod, et, transit_abs=TRUE)

par(mfrow=c(1,1))
with(transit,matplot(time,cen, type="l", ylab="Central Concentration", xlab=""))

```


Another option is to specify the transit compartment function `transit` syntax.  This specifies the parameters 
`transit(number of transit compartments, mean transit time, bioavailibility)`.  The bioavailibity term is optional. 

Using the `transit` code also automatically turns on the `transit_abs`
option.  Therefore, the same model can be specified by:


```{r out.width="100%"}

mod <- RxODE({
    ## Table 3 from Savic 2007
    cl = 17.2 # (L/hr)
    vc = 45.1 # L
    ka = 0.38 # 1/hr
    mtt = 0.37 # hr
    bio=1
    n = 20.1
    k = cl/vc
    ktr = (n+1)/mtt
    d/dt(depot) = transit(n,mtt,bio)-ka*depot
    d/dt(cen) = ka*depot-k*cen
})

et <- eventTable();
et$add.sampling(seq(0, 7, length.out=200));
et$add.dosing(20, start.time=0);

transit <- rxSolve(mod, et)

par(mfrow=c(1,1))
with(transit,matplot(time,cen, type="l", ylab="Central Concentration", xlab=""))
```


#### Stiff ODEs with Jacobian Specification

Occasionally, you may come across
a
[**stiff** differential equation](https://en.wikipedia.org/wiki/Stiff_equation),
that is a differential equation that is numerically unstable and small
variations in parameters cause different solutions to the ODEs.  One
way to tackle this is to choose a stiff-solver, or hybrid stiff solver
(like the default LSODA).  Typically this is enough. However exact
Jacobian solutions may increase the stability of the ODE.  (Note the
Jacobian is the derivative of the ODE specification with respect to
each variable). In RxODE you can specify the Jacobian with the
`df(state)/dy(variable)=` statement.  A classic ODE that has stiff
properties under various conditions is
the
[van dar Pol ](http://www.ece.northwestern.edu/local-apps/matlabhelp/techdoc/math_anal/diffeq6.html) differential
equations.

In RxODE these can be specified by the following:

```{r}
Vtpol2 <- RxODE({
    d/dt(y)  = dy
    d/dt(dy) = mu*(1-y^2)*dy - y
    ## Jacobian
    df(y)/dy(dy)  = 1
    df(dy)/dy(y)  = -2*dy*mu*y - 1
    df(dy)/dy(dy) = mu*(1-y^2)
    ## Initial conditions
    y(0) = 2
    dy(0) = 0
    ## mu
    mu = 1 ## nonstiff; 10 moderately stiff; 1000 stiff
})

et <- eventTable();
et$add.sampling(seq(0, 10, length.out=200));
et$add.dosing(20, start.time=0);

(s1 <- Vtpol2 %>%  solve(et))
```

While this is not stiff at mu=1, mu=1000 is a stiff system

```{r}
(s2 <- Vtpol2 %>%  solve(c(mu=1000), et))
```

While this is easy enough to do, it is a bit tedious.  If you have
RxODE setup appropriately, that is you have:

- **Python** installed in your system
- **sympy** installed in your system
- **SnakeCharmR** installed in R

You can use the computer algebra system sympy to calculate the
Jacobian automatically.

This is done by the RxODE option `calcJac` option:

```
Vtpol <- RxODE({
    d/dt(y)  = dy
    d/dt(dy) = mu*(1-y^2)*dy - y
    ## Initial conditions
    y(0) = 2
    dy(0) = 0
    ## mu
    mu = 1 ## nonstiff; 10 moderately stiff; 1000 stiff
}, calcJac=TRUE)

```
To see the generated model, you can use `rxCat`:

```
> rxCat(Vtpol)
d/dt(y)=dy;
d/dt(dy)=mu*(1-y^2)*dy-y;
y(0)=2;
dy(0)=0;
mu=1;
df(y)/dy(y)=0;
df(dy)/dy(y)=-2*dy*mu*y-1;
df(y)/dy(dy)=1;
df(dy)/dy(dy)=mu*(-Rx_pow_di(y,2)+1);

```



#### Simulation of Variability with RxODE
Variability in model parameters can be simulated by creating a matrix
of parameter values for use in the simulation. In the example below,
40% variability in clearance is simulated.

```{r}

mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2;
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
})

theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
               Q=1.05E+01, V3=2.97E+02,                # peripheral
               Kin=1, Kout=1, EC50=200)                # effects  
```

Each subproblem can be simulated by using the rxSolve function to run
the simulation for each set of parameters of in the parameter matrix.

```{r fig.width=10}
## the column names of the omega matrix need to match the parameters specified by RxODE
omega <- matrix(0.4^2,dimnames=list(NULL,c("eta.Cl")))

ev <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=100));

sim  <- rxSolve(mod,theta,ev,omega=omega,nSub=100)

library(ggplot2)
library(gridExtra)

p1 <- ggplot(sim,aes(time,centr,color=factor(sim.id))) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)") + guides(color=FALSE)

p2 <-ggplot(sim,aes(time,eff,color=factor(sim.id))) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

It is now straightforward to perform calculations and generate plots
with the simulated data. Below,  the 5th, 50th, and 95th percentiles
of the simulated data are plotted. 

```{r fig.width=10}

library(dplyr)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$eff, probs=p), 
                  eff.n = length(.$eff), eff.avg = mean(.$eff),
                  centr=quantile(.$centr, probs=p),
                  centr.n=length(.$centr),centr.avg = mean(.$centr))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```


Note that you can see the parameters that were simulated for the example

```{r}
head(sim$param)
```

You can also supply a data-frame of parameters to simulate instead of
using an omega simulation.  In this contrived example we will use the
previously simulated data.

```{r}
theta <- sim$param;
(sim  <- rxSolve(mod,theta,ev))
```

Even though multiple subjects were simulated, this is still a reactive
data frame, meaning you can change things about the model on the fly.

For example, if the effect at time 0 should have been 100, you can fix
this by:

```{r}
sim$eff0 <- 100
sim
```

#### Simulation of unexplained variability 

In addition to conveniently simulating between subject variability,
you can also easily simulate unexplained variability.

```{r}
mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2;
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
    e = eff+eff.err
    cp = centr*(1+cp.err)
})

theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
           Q=1.05E+01, V3=2.97E+02,                # peripheral
           Kin=1, Kout=1, EC50=200)                # effects  

sigma <- diag(2)*0.1
dimnames(sigma) <- list(NULL, c("eff.err","cp.err"))


sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```
#### Simulation of Individuals

Sometimes you may want to match the dosing and observations of
individuals in a clinical trial.  To do this you will have to create a
data.frame using the `RxODE` event specification as well as an `ID`
column to indicate an individual. The RxODE event vignette talks more about
how these datasets should be created.

If you have a NONMEM/Monlix dataset and the package
`nlmixr`, you can convert the `NONMEM`` dataset to a `RxODE` compatible dataset to
use for simulation with the `nmDataConvert` function.

Instead of using nlmixr for this simple example, I will combine two
RxODE event tables.

```{r}

ev1 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=10));

ev2 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=5000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=8));

dat <- rbind(data.frame(ID=1, ev1$get.EventTable()),
             data.frame(ID=2, ev2$get.EventTable()))


## Note the number of subject is not needed since it is determined by the data
sim  <- rxSolve(mod, theta, dat, omega=omega, sigma=sigma)

sim %>% select(id, time, e, cp)
```

#### Simulation of Clinical Trials

By either using a simple single event table, or data from a clinical
trial as described above, a complete clinical trial simulation can be
performed.

Typically in clinical trial simulations you want to account for the
uncertainty in the fixed parameter estimates, and even the uncertainty
in both your between subject variability as well as the unexplained
variability.

`RxODE` allows you to account for these uncertainties by simulating
multiple virtual "studies," specified by the parameter `nStud`.  In a
single virtual study:

- A Population effect parameter is sampled from a multivariate normal
  distribution with mean given by the parameter estimates and the
  variance specified by the named matrix `thetaMat`.
  
- A between subject variability/covariance matrix is sampled from
  either a scaled inverse chi-squared distribution (for the univariate
  case) or a inverse Wishart that is parameterized to scale to the
  conjugate prior covariance term, as described by
  the
  [wikipedia article](https://en.wikipedia.org/wiki/Scaled_inverse_chi-squared_distribution). (This
  is not the same as
  the
  [scaled inverse Wishart distribution](http://andrewgelman.com/2012/08/22/the-scaled-inverse-wishart-prior-distribution-for-a-covariance-matrix-in-a-hierarchical-model/) ).
  In the case of the between subject variability, the
  variance/covariance matrix is given by the 'omega' matrix parameter
  and the degrees of freedom is the number of subjects in the
  simulation.
  
- Unexplained variability is also simulated from the scaled inverse
  chi squared distribution or inverse Wishart distribution with the
  variance/covariance matrix given by the 'sigma' matrix parameter and
  the degrees of freedom given by the number of observations being
  simulated.
  
The covariance/variance prior is simulated from `RxODE`s `cvPost` function.

An example of this simulation is below:

```{r}

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(NULL, names(theta))

sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10)

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))

p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)

```
If you wish you can see what `omega` and `sigma` was used for each
virtual study by accessing them in the solved data object with
`$omega.list` and `$sigma.list`:

```{r}
sim$omega.list
```

```{r}
sim$sigma.list
```

You can also see the parameter realizations from the `$params` data frame.

If you do not wish to sample from the prior distributions of either
the `omega` or `sigma` matrices, you can turn off this feature by
specifying the `simVariability = FALSE` option when solving:

```{r}
sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                simVariability=FALSE);

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))


p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

Note since realizations of `omega` and `sigma` were not simulated, `$omega.list` and
`$sigma.list` both return `NULL`.

#### RxODE multi-threaded solving and simulation

RxODE now supports multi-threaded solving on OpenMP supported
compilers, including linux and windows. Mac OSX can also be supported
but
takes
[additional setup](https://github.com/Rdatatable/data.table/wiki/Installation#openmp-enabled-compiler-for-mac).
By default it uses all your available cores for solving as determined
by `rxCores()`.  This may be overkill depending on your system, at a
certain point the speed of solving is limited by things other than
computing power.

You can also speed up simulation by using the multi-cores to generate
random deviates with `mvnfast`.  This is controlled by the `nCoresRV`
parameter.  For example:

```{r}
sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                nCoresRV=rxCores());

p <- c(0.05, 0.5, 0.95);
s <-sim %>% group_by(time) %>%
    do(data.frame(p=p, eff=quantile(.$e, probs=p), 
                  eff.n = length(.$e), eff.avg = mean(.$e),
                  centr=quantile(.$cp, probs=p),
                  centr.n=length(.$cp),centr.avg = mean(.$cp))) %>%
    mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))


p1 <- ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time (hr)")

p2 <-ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time (hr)") + guides(color=FALSE)

grid.arrange(p1,p2,nrow=2)
```

The default for this is `1` core since the result depends on the
number of cores and the random seed you use in your simulation.
However, you can always speed up this process with more cores if you
are sure your collaborators have the same number of cores available to
them and have OpenMP thread-capable compile.


#### Facilities for generating R shiny applications

An example of creating an
R [shiny application](http://shiny.rstudio.com) to interactively
explore responses of various complex dosing regimens is available at
http://qsp.engr.uga.edu:3838/RxODE/RegimenSimulator.  Shiny
applications like this one may be programmatically created with the
experimental function `genShinyApp.template()`.

The above application includes widgets for varying the dose, dosing
regimen, dose cycle, and number of cycles.

```

genShinyApp.template(appDir = "shinyExample", verbose=TRUE)

library(shiny)
runApp("shinyExample")

```

### [Click here to go to the Shiny App](http://qsp.engr.uga.edu:3838/RxODE/RegimenSimulator)

```{r cleanup, include = FALSE}
mod1$dynUnload()   # otherwise we can't unlink(work)
unlink(work, recursive = TRUE)
```
