---
title: "RxODE Simulation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RxODE Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
options(knitr.table.format = "html")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')
```

# Population Simulations with RxODE

## Simulation of Variability with RxODE

In pharmacometrics the nonlinear-mixed effect modeling software (like
nlmixr) characterizes the between-subject varaibility.  With this
between subject varibility you can simulate new subjects.

Assuming that you have a 2-compartment, indirect response model, you
can set create an RxODE model describing this system below:

### Setting up the RxODE model

```{r}
library(RxODE)
mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2*(1+prop.err);
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
})
```

### Adding  the parameter estimates

The next step is to get the parameters into R so that you can start the simulation:

```{r}
theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
           Q=1.05E+01, V3=2.97E+02,                # peripheral
           Kin=1, Kout=1, EC50=200, prop.err=0)      # effects
```

In this case, I use `lotri` to specify the omega since it uses similar
lower-triangular matrix specification as nlmixr (also similar to
NONMEM):

```{r}
## the column names of the omega matrix need to match the parameters specified by RxODE
omega <- lotri(eta.Cl ~ 0.4^2)
omega
```

### Simulating 

The next step to simulate is to create the dosing regimen for overall simulation:

```{r}
ev <- et(amount.units="mg", time.units="hours") %>%
    et(amt=10000, cmt="centr")
```
If you wish, you can also add sampling times (though now RxODE can fill these in for you):

```{r}
ev <- ev %>% et(0,48, length.out=100);
```

Note the `et` takes similar arguments as `seq` when adding sampling
times.  There are more methods to adding sampling times and events to
make complex dosing regimens (See [the event
vignette](RxODE-events.html)). This includes ways to add variability
to the [both the sampling and dosing
times](RxODE-events.html#add-doses-and-samples-within-a-sampling-window)).

Once this is complete you can simulate using the `rxSolve` routine:

```{r}
if (file.exists(system.file("sim-var-sim.rds", package = "RxODE"))){
    load(system.file("sim-var-sim.rds", package = "RxODE"))
} else {
    sim  <- rxSolve(mod,theta,ev,omega=omega,nSub=100)
    save(sim, file = file.path(system.file(package = "RxODE"), "sim-var-sim.rds"))
}
```

To quickly look and customize your simulation you use the default
`plot` routine.  Since this is an RxODE object, it will create a
`ggplot2` object that you can modify as you wish.  The extra parameter
to the `plot` tells `RxODE`/`R` what piece of information you are
interested in plotting. In this case, we are interested in looking at
the derived parameter `C2`:

### Checking the simulation by a simple ggplot 

```{r}
library(ggplot2)

plot(sim, C2) +
    coord_trans(y = "log10") +
    ylab("Central Compartment") +
    xlab("Time") +
    guides(color=FALSE)
```

Of course this additional parameter could also be a state value, like `eff`:

```{r}
plot(sim, "eff") +
    coord_trans(y = "log10") +
    ylab("Effect") +
    xlab("Time") +
    guides(color=FALSE)
```

Or you could even look at the two side-by-side:

```{r}
plot(sim, C2, eff) +
    coord_trans(y = "log10") +
    ylab("Effect") +
    xlab("Time") +
    guides(color=FALSE)
```

### Processing the data to create summary plots

Usually in pharmacometric simulations it is not enough to simply
simulate the system.  We have to do something easier to digest, like
look at the central and extreme tendencies of the simulation.

Since the `RxODE` solve object is a type of [data
frame](RxODE-data-frame.html)

It is now straightforward to perform calculations and generate plots
with the simulated data. Below,  the 5th, 50th, and 95th percentiles
of the simulated data are plotted. 

```{r}
library(dplyr)
library(ggplot2)

if (file.exists("sim-var-s.rds")){
    load(system.file("sim-var-s.rds", package = "RxODE"))
} else {
    p <- c(0.05, 0.5, 0.95);
    s <-sim %>% group_by(time) %>%
        do(data.frame(p=p, eff=quantile(.$eff, probs=p), 
                      eff.n = length(.$eff), eff.avg = mean(.$eff),
                      centr=quantile(.$centr, probs=p),
                      centr.n=length(.$centr),centr.avg = mean(.$centr))) %>%
        mutate(Percentile=factor(sprintf("%d%%",p*100),levels=c("5%","50%","95%")))
    save(s, file = file.path(system.file(package = "RxODE"), "sim-var-s.rds"))
}


ggplot(s,aes(time,centr,color=Percentile)) + geom_line(size=1) + coord_trans(y = "log10") + ylab("Central Concentration") +
    xlab("Time")
```

```{r}
ggplot(s,aes(time,eff,color=Percentile)) + geom_line(size=1) + ylab("Effect") +
    xlab("Time") + guides(color=FALSE)

```

Note that you can see the parameters that were simulated for the example

```{r}
head(sim$param)
```

### Simulation of unexplained variability 

In addition to conveniently simulating between subject variability,
you can also easily simulate unexplained variability.

```{r}
mod <- RxODE({
    eff(0) = 1
    C2 = centr/V2;
    C3 = peri/V3;
    CL =  TCl*exp(eta.Cl) ## This is coded as a variable in the model
    d/dt(depot) =-KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri)  =                    Q*C2 - Q*C3;
    d/dt(eff)  = Kin - Kout*(1-C2/(EC50+C2))*eff;
    e = eff+eff.err
    cp = centr*(1+cp.err)
})

theta <- c(KA=2.94E-01, TCl=1.86E+01, V2=4.02E+01,  # central 
           Q=1.05E+01, V3=2.97E+02,                # peripheral
           Kin=1, Kout=1, EC50=200)                # effects  

sigma <- lotri(eff.err ~ 0.1, cp.err ~ 0.1)


if (file.exists(system.file("sim-var-s1.rds", package = "RxODE"))){
    load(system.file("sim-var-s1.rds", package = "RxODE"))
} else {
    sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma)
    s <- confint(sim, c("eff", "centr"));
    save(s, file = file.path(system.file(package = "RxODE"), "sim-var-s1.rds"));
}

plot(s)
```

### Simulation of Individuals

Sometimes you may want to match the dosing and observations of
individuals in a clinical trial.  To do this you will have to create a
data.frame using the `RxODE` event specification as well as an `ID`
column to indicate an individual. The RxODE event vignette talks more about
how these datasets should be created.

If you have a NONMEM/Monlix dataset and the package
`nlmixr`, you can convert the `NONMEM`` dataset to a `RxODE` compatible dataset to
use for simulation with the `nmDataConvert` function.

Instead of using nlmixr for this simple example, I will combine two
RxODE event tables.

```{r}

ev1 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=10000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=10));

ev2 <- eventTable(amount.units="mg", time.units="hours") %>%
    add.dosing(dose=5000, nbr.doses=1, dosing.to=2) %>%
    add.sampling(seq(0,48,length.out=8));

dat <- rbind(data.frame(ID=1, ev1$get.EventTable()),
             data.frame(ID=2, ev2$get.EventTable()))


## Note the number of subject is not needed since it is determined by the data
sim  <- rxSolve(mod, theta, dat, omega=omega, sigma=sigma)

sim %>% select(id, time, e, cp)
```

#### Simulation of Clinical Trials

By either using a simple single event table, or data from a clinical
trial as described above, a complete clinical trial simulation can be
performed.

Typically in clinical trial simulations you want to account for the
uncertainty in the fixed parameter estimates, and even the uncertainty
in both your between subject variability as well as the unexplained
variability.

`RxODE` allows you to account for these uncertainties by simulating
multiple virtual "studies," specified by the parameter `nStud`.  In a
single virtual study:

- A Population effect parameter is sampled from a multivariate normal
  distribution with mean given by the parameter estimates and the
  variance specified by the named matrix `thetaMat`.
  
- A between subject variability/covariance matrix is sampled from
  either a scaled inverse chi-squared distribution (for the univariate
  case) or a inverse Wishart that is parameterized to scale to the
  conjugate prior covariance term, as described by
  the
  [wikipedia article](https://en.wikipedia.org/wiki/Scaled_inverse_chi-squared_distribution). (This
  is not the same as
  the
  [scaled inverse Wishart distribution](http://andrewgelman.com/2012/08/22/the-scaled-inverse-wishart-prior-distribution-for-a-covariance-matrix-in-a-hierarchical-model/) ).
  In the case of the between subject variability, the
  variance/covariance matrix is given by the 'omega' matrix parameter
  and the degrees of freedom is the number of subjects in the
  simulation.
  
- Unexplained variability is also simulated from the scaled inverse
  chi squared distribution or inverse Wishart distribution with the
  variance/covariance matrix given by the 'sigma' matrix parameter and
  the degrees of freedom given by the number of observations being
  simulated.
  
The covariance/variance prior is simulated from `RxODE`s `cvPost()` function.

An example of this simulation is below:

```{r}

## Creating covariance matrix
tmp <- matrix(rnorm(8^2), 8, 8)
tMat <- tcrossprod(tmp, tmp) / (8 ^ 2)
dimnames(tMat) <- list(NULL, names(theta))

if (file.exists(system.file("sim-var-s2.rds", package = "RxODE"))){
    load(system.file("sim-var-s2.rds", package = "RxODE"))
} else {
    sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                    dfSub=10, dfObs=100)

    s <-sim %>% confint(c("centr", "eff"))
    save(sim, s, file = file.path(system.file(package = "RxODE"), "sim-var-s2.rds"));
}
plot(s)
```

If you wish you can see what `omega` and `sigma` was used for each
virtual study by accessing them in the solved data object with
`$omega.list` and `$sigma.list`:

```{r}
head(sim$omega.list)
```

```{r}
head(sim$sigma.list)
```

You can also see the parameter realizations from the `$params` data frame.

If you do not wish to sample from the prior distributions of either
the `omega` or `sigma` matrices, you can turn off this feature by
specifying the `simVariability = FALSE` option when solving:

```{r}

if (file.exists(system.file("sim-var-s3.rds", package = "RxODE"))){
    load(system.file("sim-var-s3.rds", package = "RxODE"))
} else {

    sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                    simVariability=FALSE);

    s <-sim %>% confint(c("centr", "eff"))
    save(sim, s, file = file.path(system.file(package = "RxODE"), "sim-var-s3.rds") );

}


plot(s)
```


Note since realizations of `omega` and `sigma` were not simulated, `$omega.list` and
`$sigma.list` both return `NULL`.

#### RxODE multi-threaded solving and simulation

RxODE now supports multi-threaded solving on OpenMP supported
compilers, including linux and windows. Mac OSX can also be supported
but
takes
[additional setup](https://github.com/Rdatatable/data.table/wiki/Installation#openmp-enabled-compiler-for-mac).
By default it uses all your available cores for solving as determined
by `rxCores()`.  This may be overkill depending on your system, at a
certain point the speed of solving is limited by things other than
computing power.

You can also speed up simulation by using the multi-cores to generate
random deviates with `mvnfast` (either `mvnfast::rmvn()` or
`mvnfast::rmvt()`).  This is controlled by the `nCoresRV` parameter.
For example:

```{r}
if (file.exists(system.file("sim-var-s4.rds", package = "RxODE"))){
    load(system.file("sim-var-s4.rds", package = "RxODE"))
} else {

    sim  <- rxSolve(mod, theta, ev, omega=omega, nSub=100, sigma=sigma, thetaMat=tMat, nStud=10,
                    nCoresRV=2);

    p <- c(0.05, 0.5, 0.95);
    s <-sim %>% confint(c("eff", "centr"))
    save(sim, s, file = file.path(system.file(package = "RxODE"), "sim-var-s4.rds"))
}

plot(s)
```

The default for this is `1` core since the result depends on the
number of cores and the random seed you use in your simulation.
However, you can always speed up this process with more cores if you
are sure your collaborators have the same number of cores available to
them and have OpenMP thread-capable compile.

