---
title: "Speeding up RxODE"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
options(knitr.table.format = "html")
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'RxODE', 
               style = 'position:absolute; top:0; right:0; padding:10px; border: 0;')

```

# Increasing RxODE speed by multi-subject parallel solving
`RxODE` originally developed as an ODE solver that allowed an ODE
solve for a single subject.  This flexibility is still supported.

The original code from the `RxODE` tutorial is below:

```{r}

library(RxODE)
library(microbenchmark)
library(mvnfast)

mod1 <- RxODE({
    C2 = centr/V2;
    C3 = peri/V3;
    d/dt(depot) = -KA*depot;
    d/dt(centr) = KA*depot - CL*C2 - Q*C2 + Q*C3;
    d/dt(peri) = Q*C2 - Q*C3;
    d/dt(eff) = Kin - Kout*(1-C2/(EC50+C2))*eff;
    eff(0) = 1
})

## Create an event table

ev <- et() %>%
    et(amt=10000, addl=9,ii=12) %>%
    et(time=120, amt=20000, addl=4, ii=24) %>%
    et(0:240) ## Add Sampling

nsub <- 100 # 100 subproblems
sigma <- matrix(c(0.09,0.08,0.08,0.25),2,2) # IIV covariance matrix
mv <- rmvn(n=nsub, rep(0,2), sigma) # Sample from covariance matrix
CL <- 7*exp(mv[,1])
V2 <- 40*exp(mv[,2])
params.all <- cbind(KA=0.3, CL=CL, V2=V2, Q=10, V3=300,
                    Kin=0.2, Kout=0.2, EC50=8)
```

## For Loop
The slowest way to code this is to use a `for` loop.  In this example
we will enclose it in a function to compare timing.

```{r}
runFor <- function(){
    res <- NULL
    for (i in 1:nsub) {
        params <- params.all[i,]
        x <- mod1$solve(params, ev)
        ##Store results for effect compartment
        res <- cbind(res, x[, "eff"])
    }
    return(res)
}
```
## Running with apply 

In general for R, the `apply` types of functions perform better than a
`for` loop, so the tutorial also suggests this speed enhancement

```{r}
runSapply <- function(){
    res <- apply(params.all, 1, function(theta)
        mod1$run(theta, ev)[, "eff"])
}
```
## Run using a single-threaded solve

You can also have RxODE solve all the subject simultaneously without
collecting the results in R, using a single threaded solve.

The data output is slightly different here, but still gives the same
information:

```{r}
runSingleThread <- function(){
    solve(mod1, params.all, ev, cores=1)[,c("sim.id", "time", "eff")]
}
```

## Run a 2 threaded solve

RxODE supports multi-threaded solves, so another option is to have `2`
threads (called `cores` in the solve options, you can see the options
in `rxControl()` or `rxSolve()`). 

```{r}
run2Thread <- function(){
    solve(mod1, params.all, ev, cores=2)[,c("sim.id", "time", "eff")]
}
```

## Compare the times between all the methods

Now the moment of truth, the timings:

```{r}
if (file.exists("benchAll.rds")){
    load(file="benchAll.rds");
} else {
    bench <- microbenchmark(runFor(), runSapply(), runSingleThread(),run2Thread())
    save(bench,file="benchAll.rds");
}
print(bench)
```

```{r}
plot(bench)
```

It is clear that the **largest** jump in performance when using the
`solve` method and providing *all* the parameters to RxODE to solve
without looping over each subject with either a `for` or a `sapply`.
The number of cores/threads applied to the solve also plays a role in
the solving.

We can explore the number of threads further with the following code:

```{r}
runThread <- function(n){
    solve(mod1, params.all, ev, cores=n)[,c("sim.id", "time", "eff")]
}

if (file.exists("benchThread.rds")){
    load(file="benchThread.rds");
} else {
    bench <- eval(parse(text=sprintf("microbenchmark(%s)",
                                     paste(paste0("runThread(", 1:12,")"),
                                           collapse=","))))
    save(bench,file="benchThread.rds");
}

print(bench)
```

```{r}
plot(bench)
```

In this case, the best number of threads to choose is 4 threads.  The
complexity of the ODE solving and the number of subjects may affect
this arbitrary number of threads.  At the same time 4 is a good number
to use without any prior knowledge because most systems these days
have at least 4 threads (or 2 processors with 4 threads).

## Want more ways to run multi-subject simulations
The version since the tutorial has even more ways to run multi-subject
simulations, including adding variability in sampling and dosing times
with `et()` (see [RxODE events](RxODE-events.html#add-doses-and-samples-within-a-sampling-window) for more
information), ability to supply both an `omega` and `sigma` matrix as
well as adding as a `thetMat` to R to simulate with uncertainty in the
`omega`, `sigma` and `theta` matrices; see [RxODE simulation
vignette](RxODE-sim-var.html).
